---

title: "LUAD Multi-Omics Classification: Data Preprocessing and Integration"
output:
html_document:
df_print: paged
---------------

# Data Preprocessing and Integration

## Project Overview

This notebook performs data acquisition, preprocessing, quality control, and integration of transcriptomic (RNA-seq), DNA methylation, and clinical data from TCGA Lung Adenocarcinoma (LUAD) samples.

The objective is to generate a harmonized multi-omics dataset suitable for downstream differential analysis, feature selection, and machine learning classification.

## Pipeline Position

Step 1 of 4

Data Acquisition → Data Integration → Feature Selection & Modeling → Evaluation → Biological Interpretation

## Inputs

* TCGA LUAD RNA-seq data
* TCGA LUAD DNA methylation data
* Clinical metadata

## Outputs

* Matched multi-omics sample set
* Filtered expression matrix
* Filtered methylation matrix
* Stratified training and test datasets

## Authors

Developed as part of a collaborative academic project at Freie Universität Berlin.


# Section 1: Data and Integration Steps

Our final project focused on working with the TCGA LUAD (Lung Adenocarcinoma) RNA-Seq, Methylation and Clinical data. This R Notebook contains the code corresponding to the following steps of the omics data analysis cycle:

1.  **Data**

    a\. Download (RNA-Seq, Clinical Data & Methylation)

    b\. Exploration

    c\. Preprocessing

2.  **Integration**

    a\. Quality Control

    b\. Filtering

    c\. Sample Matching between Omics Datasets

    d\. Stratified Split into Train-Test Datasets

    e\. Data Subsetting for Downstream Pairwise Comparisons

# 1.1 Setup

Load required libraries.

```{r setup, message=FALSE, warning=FALSE}

# Global chunk options
knitr::opts_chunk$set(echo = TRUE)

# Install Bioconductor only if not already installed
#if (!requireNamespace("BiocManager", quietly = TRUE))
    #install.packages("BiocManager")

# Bioconductor Packages
#BiocManager::install(c("recount3", "SummarizedExperiment", "DESeq2", "TCGAutils", "TCGAbiolinks", "curatedTCGAData", "sesame", "sesameData", "org.Hs.eg.db"))

# CRAN Packages
#install.packages("tidyverse", "caret", "dplyr", "pheatmap", "tibble", "ggplot2", "reshape2")

# Load libraries
library(recount3) #RNA-Seq data
library(SummarizedExperiment) #RNA-seq data manipulation
library(tidyverse) # Data Manipulation
library(DESeq2) #DEG
library(caret) # Stratification
library(dplyr) # Stratification
library(curatedTCGAData) #Cancer subtypes and stage
library(TCGAbiolinks) # Methylation data
library(sesame) #Methylation
library(sesameData) #Methylation
library(org.Hs.eg.db) #Map to HGNC gene symbols

# Visualization
library(pheatmap)
library(tibble)
library(ggplot2)
library(reshape2)
library(VennDiagram)
library(grid)

```

Set working directory. Intermediate objects will be saved in and loaded from folders created in this working location.

```{r work-dir-setup}

work_dir <- "/home/rstudio/data/FinalProject/LUAD Project RData/Final_Code/"

# Set working directory
setwd(work_dir)

```

# 1.2 RNA-Seq Data

## 1.2.1 Download

RNA-seq expression data download via the *recount3* package.

```{r rna-data-download, message=FALSE}

# Available human projects in recount3
human_projects <- available_projects()

# Filter only TCGA LUAD projects
tcga_luad_project <- subset(human_projects, project == "LUAD" & file_source == "tcga")

# Display quick view of TCGA LUAD project information
tcga_luad_project

```

RangeSummarizedExperiment (RSE) object creation, this object contains raw counts and metadata.

```{r rna-data-rse, message=FALSE}

# Create RangedSummarizedExperiment (RSE)
rse_tcga_luad <- create_rse(tcga_luad_project)

# Display summary of TCGA LUAD RSE object information
rse_tcga_luad

```

## 1.2.2 Clinical Data Download- Subtype Information

Clinical data download via the *curatedTCGAData* package.

```{r clin-data-download, message=FALSE}

luad_curated <- curatedTCGAData(
  "LUAD",
  assays = "RNASeq2GeneNorm",
  dry.run = FALSE,
  version = "2.1.1"
)

```

Match samples between RNA-Seq and Clinical Data to keep common sample set.

```{r sample-match-rna-clinical}

# Integrating LUAD cancer subtype data from curatedTCGAData

# Get patientID from barcode from recount3 data and clinical data (both in type character)
patient_ids_rse <- substr(rse_tcga_luad$tcga.tcga_barcode, 1, 12)
patient_ids_curated <- luad_curated$patientID 

# Use name vectors to use the patient ids
expression_subtype <- setNames(as.character(luad_curated$expression_subtype), patient_ids_curated)

# Add expression subtypes to colData from recount3, rse_tcga_luad
colData(rse_tcga_luad)$expression_subtype <- expression_subtype[patient_ids_rse]

# Make subtypes factor for downstream analysis
colData(rse_tcga_luad)$expression_subtype <- factor(
  colData(rse_tcga_luad)$expression_subtype,
  levels = c("prox.-inflam", "prox.-prolif.", "TRU")
)

# Add sample type as factor column
colData(rse_tcga_luad)$sample_type_factor <- factor(
  rse_tcga_luad$tcga.cgc_sample_sample_type,
  levels = c("Primary Tumor", "Solid Tissue Normal")
)

# Substitute spaces by underscores in levels to avoid downstream errors
levels(colData(rse_tcga_luad)$sample_type_factor) <- gsub(" ", "_", levels(colData(rse_tcga_luad)$sample_type_factor))

# Keep only samples with sample type info to start
keep_samples <- !is.na(colData(rse_tcga_luad)$sample_type_factor)
rse_sample_filtered <- rse_tcga_luad[, keep_samples]

```

Pathologic stage collapsing to obtain more general categories.

```{r stage-collapse}

# Extract raw stage
stage_raw <- rse_sample_filtered$tcga.cgc_case_pathologic_stage

# Collapse stages
stage_collapsed <- case_when(
  stage_raw %in% c("Stage I", "Stage IA", "Stage IB") ~ "Stage I",
  stage_raw %in% c("Stage II", "Stage IIA", "Stage IIB")  ~ "Stage II",
  stage_raw %in% c("Stage III", "Stage IIIA", "Stage IIIB", "Stage IIIC") ~ "Stage III",
  stage_raw == "Stage IV" ~ "Stage IV",
  TRUE ~ NA_character_ # for any rows that do not match the other conditions, unknowns X or Tis
)

# Add to rse metadata as factor
colData(rse_sample_filtered)$stage_collapsed <- factor(stage_collapsed,
    levels = c("Stage I", "Stage II", "Stage III", "Stage IV"))

# Substitute spaces by underscores in levels to avoid downstream errors
levels(colData(rse_sample_filtered)$stage_collapsed) <- gsub(" ", "_", levels(colData(rse_sample_filtered)$stage_collapsed))

```

## 1.2.3 Data Exploration

### Exploratory PCA

Exploratory PCA to show the global structure of our raw data. Raw data here refers to the RNA-Seq + Subtype information dataset.

#### RNA-Seq Data Normalization for PCA

Data normalized using the variance stabilizing transformation function *vst* from *DESeq2*.

```{r rna-raw-data-norm, message=FALSE}

# To do exploratory PCA, raw expression counts need to be normalized
# Normalize counts with vst from DESeq2

# DESeq2 object creation
dds_raw <- DESeqDataSetFromMatrix(
    countData = assays(rse_sample_filtered)$raw_counts,
    colData = colData(rse_sample_filtered),
    design = ~ sample_type_factor
)

# vst normalization
vsd_raw <- vst(dds_raw, blind = TRUE)

```

Exploratory sample type PCA visualisation.

```{r rna-raw-data-pca-sample, message=FALSE}

# PCA colored by sample, normal vs tumor
plotPCA(vsd_raw, intgroup = "sample_type_factor") +
  ggtitle("LUAD Raw Data PCA \n Colored by Sample Type") + 
  theme(plot.title = element_text(hjust = 0.5))

```

Exploratory subtype PCA visualisation.

```{r rna-raw-data-pca-subtype, message=FALSE}

# PCA colored by cancer subtype
plotPCA(vsd_raw, intgroup = "expression_subtype") +
  ggtitle("LUAD Raw Data PCA \n Colored by Subtype") + 
  theme(plot.title = element_text(hjust = 0.5))

```

Exploratory stage PCA visualisation.

```{r rna-raw-data-pca-stage, message=FALSE}

# PCA colored by stage
plotPCA(vsd_raw, intgroup = "stage_collapsed") +
  ggtitle("LUAD Raw Data PCA \n Colored by Collapsed Cancer Stage") + 
  theme(plot.title = element_text(hjust = 0.5))

```

#### Gene Types and Gene Counts

Counts per gene, check whether any genes need to be filtered out due to low counts (noise). Visualise filter suggested by *DESeq2* authors.

```{r rna-raw-genes-hist, message=FALSE, warning=FALSE}

# Histogram of counts per gene
# Sum of counts across all samples (raw, not normalized)
gene_counts_raw <- rowSums(assay(rse_sample_filtered))

ggplot(data.frame(total_counts = gene_counts_raw), aes(x = total_counts)) +
  geom_histogram(bins = 100, fill = "#2c7fb8", color = "white") +
  scale_x_log10() +
  geom_vline(xintercept = 100, linetype = "dashed", color = "red") +
  xlab("Total Counts per Gene (log10 scale)") +
  ylab("Number of Genes") +
  ggtitle("Distribution of Total Counts per Gene (Before Filtering)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

Visualize gene types on our data.

```{r rna-raw-genes-bar, message=FALSE}

# Bar plot of gene types

# Gene types table
gene_types <- rowData(rse_sample_filtered)$gene_type
gene_type_tbl <- as.data.frame(table(gene_types)) #type and frequency

# Sort top gene types
top_types <- gene_type_tbl[order(-gene_type_tbl$Freq), ][1:15, ]  # top 10 most common

# Barplot
ggplot(top_types, aes(x = reorder(gene_types, Freq), y = Freq)) +
  geom_bar(stat = "identity", fill = "#41ab5d") +
  coord_flip() +
  xlab("Gene Type") +
  ylab("Number of Genes") +
  ggtitle("Most Common Gene Types in RNA-seq Data") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

Look into most variable genes to have an idea of how informative our genes are.

```{r rna-raw-genes-heatmap}

# Heatmap of most variable genes

# Variance of each gene (using VST-normalized data)
vsd_raw_mat <- assay(vsd_raw)
gene_vars <- apply(vsd_raw_mat, 1, var)

# Create annotation for subtypes (columns = samples)
annotation_col <- data.frame(
  Subtype = colData(vsd_raw)$expression_subtype
)
rownames(annotation_col) <- colnames(vsd_raw_mat)

# Top 30 most variable genes
top_var_genes <- head(order(gene_vars, decreasing = TRUE), 30)

# Heatmap top 30
pheatmap(
  vsd_raw_mat[top_var_genes, ],
  scale = "row",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  annotation_col = annotation_col,
  show_rownames = FALSE,
  show_colnames = FALSE,
  main = "Top 30 Most Variable Genes (VST Transformed)"
)

```

## 1.2.4 Data Preprocessing & Filtering

Preprocessing & filtering based on the data exploration observations.

### Filtering by sample type

```{r sample-filtering}

# Filter by sample type, keep only normal and primary tumor
sample_types <- colData(rse_sample_filtered)$sample_type_factor
keep_samples <- sample_types %in% c("Primary_Tumor", "Solid_Tissue_Normal") & !is.na(sample_types)
rse_luad_filtered_samples <- rse_sample_filtered[, keep_samples]

```

### Filtering by gene type

```{r gene-filtering}

# Filter by gene type, keep only protein coding genes
keep_genes <- rowData(rse_sample_filtered)$gene_type == "protein_coding"
rse_luad_filtered_genes <- rse_luad_filtered_samples[keep_genes, ]

```

### Filtering by gene counts

```{r gene-counts-filtering}

# Filter genes by expression counts as recommended by DESeq2 authors
# 10 counts in at least 10 samples

# Extract raw counts into an object to filter
luad_countData<- assays(rse_luad_filtered_genes)$raw_counts

# Keep genes expressed at least 10 counts in 10+ samples
keep_genes <- rowSums(luad_countData >= 10) >= 10

# Update rse object, not just the extracted countData
# s sample, g gene type, c gene counts
rse_luad_sgc_filtered <- rse_luad_filtered_genes[keep_genes, ]

```

## 1.2.5 Filtered Data Visualisation

### PCA after Filtering

Filtered data normalised again with *vst* from *DESeq2*

```{r rna-filtered-data-norm, message=FALSE}

# Intermediate rse objects with specific samples, for plotting PCA after filtering
rse_stage_primary <- rse_luad_sgc_filtered[
     , rse_luad_sgc_filtered$tcga.cgc_sample_sample_type == "Primary Tumor" &
         !is.na(rse_luad_sgc_filtered$stage_collapsed)
]

rse_subtype_primary <- rse_luad_sgc_filtered[
  , rse_luad_sgc_filtered$tcga.cgc_sample_sample_type == "Primary Tumor" &
        !is.na(rse_luad_sgc_filtered$expression_subtype)
]

#Creating DESeq2 objects for PCA

dds_sgc_filtered <- DESeqDataSetFromMatrix(
    countData = assays(rse_luad_sgc_filtered)$raw_counts,
    colData = colData(rse_luad_sgc_filtered),
    design = ~ sample_type_factor
)

dds_filtered_subtype <- DESeqDataSetFromMatrix(
  countData = assays(rse_subtype_primary)$raw_counts,
  colData = colData(rse_subtype_primary),
  design = ~ 1
)

dds_filtered_stage <- DESeqDataSetFromMatrix(
  countData = assays(rse_stage_primary)$raw_counts,
  colData = colData(rse_stage_primary),
  design = ~ 1
)

# Use vst to stabilize counts by variance
vsd_sgc_filtered <- vst(dds_sgc_filtered, blind = TRUE)
vsd_filtered_stage <- vst(dds_filtered_stage, blind = TRUE)
vsd_filtered_subtype <- vst(dds_filtered_subtype, blind = TRUE)

```

Filtered subtype PCA visualisation.

```{r rna-filtered-data-pca-subtype, message=FALSE}

# Filtered data colored by subtype
plotPCA(vsd_filtered_subtype, intgroup = "expression_subtype") +
  ggtitle("LUAD Filtered Data PCA \n Colored by Subtype") + 
  theme(plot.title = element_text(hjust = 0.5))

```

Filtered stage PCA visualisation.

```{r rna-filtered-data-pca-stage, message=FALSE}

# Filtered data colored by stage
plotPCA(vsd_filtered_stage, intgroup = "stage_collapsed") +
  ggtitle("LUAD Filtered Data PCA \n Colored by Cancer Stage") + 
  theme(plot.title = element_text(hjust = 0.5))

```

Gene count distribution after filtering, showing gene count filter line for reference.

```{r rna-filtered-genes-hist, message=FALSE, warning=FALSE}

# Histogram of counts per gene
# Sum of counts across all samples (raw, not normalized)
gene_counts_filtered <- rowSums(assay(rse_luad_sgc_filtered))

ggplot(data.frame(total_counts = gene_counts_filtered), aes(x = total_counts)) +
  geom_histogram(bins = 100, fill = "#2c7fb8", color = "white") +
  scale_x_log10() +
  geom_vline(xintercept = 100, linetype = "dashed", color = "red") +
  xlab("Total Counts per Gene (log10 scale)") +
  ylab("Number of Genes") +
  ggtitle("Distribution of Total Counts per Gene (After Filtering)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Heatmap of most variable genes after filtering.

```{r rna-filtered-genes-heatmap}

# Variance of each gene (using VST-normalized data)
vsd_filtered_mat <- assay(vsd_sgc_filtered)
gene_vars <- apply(vsd_filtered_mat, 1, var)

# Create annotation for subtypes (columns = samples)
annotation_col <- data.frame(
  Subtype = colData(vsd_sgc_filtered)$expression_subtype
)
rownames(annotation_col) <- colnames(vsd_sgc_filtered)

# Top 30 most variable genes
top_var_genes <- head(order(gene_vars, decreasing = TRUE), 30)

# Heatmap top 30
pheatmap(
  vsd_filtered_mat[top_var_genes, ],
  scale = "row",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  annotation_col = annotation_col,
  show_rownames = FALSE,
  show_colnames = FALSE,
  main = "Filtered Data - Top 30 Most Variable Genes (VST Transformed)"
)

```

## 1.2.6 Stratified Split into Train and Test Sets

Global split into 80% training-20% test. The stratification will be done considering sample type, cancer subtype and tumor stage to keep the proportions as in the input filtered data.

```{r stratified-split, message=FALSE, warning=FALSE}

# Stratified Split based on metadata labels
# Extract metadata into df
luad_metadata_df <- as.data.frame(colData(rse_luad_sgc_filtered))

# Copy for before split visualization
luad_metadata_before_split_df <- luad_metadata_df

# Create a composite stratification label
# e.g., "Primary_tumor_prolif.-inflam_Stage_I", "Solid_tissue_normal_TRU_Stage_II", etc.

# Create composite strata labels
luad_metadata_df$strata <- paste(
  luad_metadata_df$sample_type_factor,
  luad_metadata_df$expression_subtype,
  luad_metadata_df$stage_collapsed,
  sep = "_"
)

# Global Stratified split using composite label, test-training
set.seed(42)
luad_train_ids <- createDataPartition(luad_metadata_df$strata, p = 0.8, list = FALSE)

# Assign split labels
luad_metadata_df$split <- "test"
luad_metadata_df$split[luad_train_ids]<- "train"

# Add split info into the rse ColData metadata
colData(rse_luad_sgc_filtered)$split <- luad_metadata_df$split

cat("Sample assignment after Stratified Split:")
table(rse_luad_sgc_filtered$split)

```

Saving intermediate RSE objects that will be useful to skip previous processing steps and can be used downstream.

```{r rna-data-save}

# Create "rse_objects" folder if it does not exist
dir.create("LUAD_rse_objects", showWarnings = FALSE)

# File paths, relative to working dir
path_filtered <- file.path("LUAD_rse_objects", "rse_luad_sgc_filtered.RData")
path_train <- file.path("LUAD_rse_objects", "rse_luad_train.RData")
path_test <- file.path("LUAD_rse_objects", "rse_luad_test.RData")

# Saving filtered object with split, subtype and sample type info added as extra metadata factor cols
save(rse_luad_sgc_filtered, file = path_filtered)

# Create 80-20 train/test RSE objects
rse_luad_train <- rse_luad_sgc_filtered[, colData(rse_luad_sgc_filtered)$split == "train"]
rse_luad_test <- rse_luad_sgc_filtered[, colData(rse_luad_sgc_filtered)$split == "test"]

# Extract only train primary tumor samples
rse_luad_train_primary <- rse_luad_train[, colData(rse_luad_train)$sample_type_factor == "Primary_Tumor"]

# Save train/test objects
save(rse_luad_train, file = path_train)
save(rse_luad_test, file = path_test)
```

### Visualisation Before Split

Visualising proportions of subtype and stage samples in filtered data before stratified split.

```{r before-stratification-plot-subtype}

# Stratification breakdown plots, before and after

# Before split subtype proportions
ggplot(subset(luad_metadata_before_split_df, !is.na(expression_subtype)), 
       aes(x = expression_subtype)) +
  geom_bar(fill = "orchid") +
  labs(title = "Subtype Distribution Before Stratified Split", 
       x = "Subtype", y = "Sample Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r before-stratification-plot-stage}

# Before split stage proportions
ggplot(subset(luad_metadata_before_split_df, !is.na(stage_collapsed)), 
       aes(x = stage_collapsed)) +
  geom_bar(fill = "cadetblue3") +
  labs(title = "Stage Distribution Before Stratified Split", 
       x = "Stage", y = "Sample Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

### Visualisation after Stratified Split

Visualizing proportions of subtype and stage samples in data after stratified split.

```{r after-stratification-plot-subtype}

# After split subtype proportions
# Filter metadata to samples with subtype info
subtype_meta <- subset(luad_metadata_df, !is.na(expression_subtype))

ggplot(subtype_meta, aes(x = expression_subtype, fill = split)) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("train" = "orchid4", "test" = "plum2")) +
  labs(title = "Subtype Distribution by Split (Filtered)", x = "Subtype", y = "Sample Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r after-stratification-plot-stage}

# After split stage proportions
stage_meta <- subset(luad_metadata_df, !is.na(stage_collapsed))

ggplot(stage_meta, aes(x = stage_collapsed, fill = split)) +
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("train" = "cadetblue", "test" = "lightblue")) +
  labs(title = "Stage Distribution by Split (Filtered)", x = "Stage", y = "Sample Count") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

## 1.2.7 Pairwise Comparison Data Subsets

Subsetting data for pairwise comparisons, subtype and stage pairs. Pair subsets will be saved in an RSE object.

```{r subset-pair-rse}
# Helper function to get samples for each comparison
subset_rse <- function(rse, filter_expr) {
  meta <- as.data.frame(colData(rse))
  subset_idx <- which(eval(parse(text = filter_expr), envir = meta))
  rse[, subset_idx]
}

# Helper function to subset and save rse
subset_and_save_rse <- function(rse, condition, name, outdir = "rse_objects") {
  message("Creating: ", name)
  subset <- subset_rse(rse, condition)
  cat("    Samples:", ncol(subset), "| Genes:", nrow(subset), "\n")
  # Save with object name as in environment, useful when loading back
  assign(name, subset, envir = .GlobalEnv)
  save(list = name, file = file.path(outdir, paste0(name, ".RData")))
}

# Define subtype comparisons
subtype_comparisons <- list(
  rse_luad_pi_vs_pp = 'sample_type_factor == "Primary_Tumor" & expression_subtype %in% c("prox.-inflam", "prox.-prolif.")',
  rse_luad_pi_vs_tru = 'sample_type_factor == "Primary_Tumor" & expression_subtype %in% c("prox.-inflam", "TRU")',
  rse_luad_pp_vs_tru = 'sample_type_factor == "Primary_Tumor" & expression_subtype %in% c("prox.-prolif.", "TRU")'
)

# Define stage comparisons
stage_comparisons <- list(
  rse_luad_stage1_vs_2 = 'sample_type_factor == "Primary_Tumor" & stage_collapsed %in% c("Stage_I", "Stage_II")',
  rse_luad_stage1_vs_3 = 'sample_type_factor == "Primary_Tumor" & stage_collapsed %in% c("Stage_I", "Stage_III")',
  rse_luad_stage1_vs_4 = 'sample_type_factor == "Primary_Tumor" & stage_collapsed %in% c("Stage_I", "Stage_IV")',
  rse_luad_stage2_vs_3 = 'sample_type_factor == "Primary_Tumor" & stage_collapsed %in% c("Stage_II", "Stage_III")',
  rse_luad_stage2_vs_4 = 'sample_type_factor == "Primary_Tumor" & stage_collapsed %in% c("Stage_II", "Stage_IV")',
  rse_luad_stage3_vs_4 = 'sample_type_factor == "Primary_Tumor" & stage_collapsed %in% c("Stage_III", "Stage_IV")'
)

# Subset and save subtype comparison rse objects
for (name in names(subtype_comparisons)) {
  subset_and_save_rse(rse_luad_train, subtype_comparisons[[name]], name)
}

# Subset and save stage comparison rse objects
for (name in names(stage_comparisons)) {
  subset_and_save_rse(rse_luad_train, stage_comparisons[[name]], name)
}
```

### Pairwise Comparisons Visualisation

Heatmap to visualize samples that went into each pairwise comparison.

```{r train-primary-metadata-processing}

# Processing train primary tumor samples data to generate heatmaps

# Extract metadata
train_meta <- as.data.frame(colData(rse_luad_train_primary))

# Helper function to get counts for each pair
get_pair_counts <- function(df, variable, group1, group2, label) {
  df_filtered <- df %>%
    filter(!is.na(.data[[variable]])) %>%
    filter(.data[[variable]] %in% c(group1, group2))
  tibble(
    Comparison = label,
    Group = c(group1, group2),
    Count = c(
      sum(df_filtered[[variable]] == group1),
      sum(df_filtered[[variable]] == group2)
    )
  )
}

# Subtype comparisons counts table
subtype_counts <- bind_rows(
  get_pair_counts(train_meta, "expression_subtype", "prox.-inflam", "prox.-prolif.", "proxInflam vs proxProlif"),
  get_pair_counts(train_meta, "expression_subtype", "TRU", "prox.-prolif.", "TRU vs proxProlif"),
  get_pair_counts(train_meta, "expression_subtype", "TRU", "prox.-inflam", "TRU vs proxInflam")
)

# Stage comparisons counts table
stage_counts <- bind_rows(
  get_pair_counts(train_meta, "stage_collapsed", "Stage_I", "Stage_II", "Stage I vs II"),
  get_pair_counts(train_meta, "stage_collapsed", "Stage_I", "Stage_III", "Stage I vs III"),
  get_pair_counts(train_meta, "stage_collapsed", "Stage_I", "Stage_IV", "Stage I vs IV"),
  get_pair_counts(train_meta, "stage_collapsed", "Stage_II", "Stage_III", "Stage II vs III"),
  get_pair_counts(train_meta, "stage_collapsed", "Stage_II", "Stage_IV", "Stage II vs IV"),
  get_pair_counts(train_meta, "stage_collapsed", "Stage_III", "Stage_IV", "Stage III vs IV")
)

```

#### Subtype Comparisons Heatmap

```{r rna-pairwise-comp-heatmap-subtype}

# Subtype pairwise subset samples
ggplot(subtype_counts, aes(x = Comparison, y = Group, fill = Count)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Count), color = "black", size = 3.5) +
  scale_fill_gradient(low = "#e0f3f8", high = "#08589e") +
  theme_minimal() +
  labs(title = "Samples per Group in Subtype Comparisons",
       x = "Comparison", y = "Subtype") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5))

```

#### Stage Comparisons Heatmap

```{r rna-pairwise-comp-heatmap-stage}

# Subtype pairwise subset samples
ggplot(stage_counts, aes(x = Comparison, y = Group, fill = Count)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Count), color = "black", size = 3.5) +
  scale_fill_gradient(low = "#fee0d2", high = "#de2d26") +
  theme_minimal() +
  labs(title = "Samples per Group in Stage Comparisons",
       x = "Comparison", y = "Stage") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5))

```

# 1.3 Methylation Data

Methylation data β values downloaded via the *TCGAbiolinks* package.

**NOTE:** Methylation data file download & consolidation takes about 1 hour to complete so we offer two options to proceed:\
\
**Option A**: Full download and consolidation of the downloaded files into a single SummarizedExperiment object (LUAD_methylation_SE.rda file), used to continue the analysis.\
\
**Option B**: For practicality, we provide the LUAD_methylation_SE.rda file that was processed through the Option A code. If needed due to time or memory constraints, the {met-data-download} and {met-data-se} code chunks can be skipped and substituted by the {met-data-se-load}, to load the provided file.

## 1.3.1 Option A: Download & SE Consolidation

```{r met-data-download}

# Define download directory
download_dir <- "LUAD_met_data_files"
if (!dir.exists(download_dir)) dir.create(download_dir)

# Build the query
luad_query_met <- GDCquery(
  project = "TCGA-LUAD",
  data.category = "DNA Methylation",
  data.type = "Methylation Beta Value",
  platform  = "Illumina Human Methylation 450"
)

# Download in batches, each time into download_dir
files <- getResults(luad_query_met)
total_files <- nrow(files)
batch_size <- 50

for (start in seq(1, total_files, by = batch_size)) {
  end <- min(start + batch_size - 1, total_files)
  batch_query <- luad_query_met
  batch_query$results[[1]] <- files[start:end, ]

  cat(sprintf("Downloading batch %d to %d...\n", start, end))
  
  tryCatch({
    GDCdownload(batch_query, directory = download_dir, method = "api")

    # Check that the files exist after download
    batch_files <- getResults(batch_query)$file_name
    missing <- !file.exists(file.path(download_dir, batch_files))

    if (any(missing)) {
      cat("Missing files in batch, re-trying...\n")
      GDCdownload(batch_query, directory = download_dir, method = "api")
    }
    
  }, error = function(e) {
    message(sprintf("Failed batch %d–%d: %s", start, end, e$message))
  })
}

```

Methylation data consolidation into SummarizedExperiment object. Memory intensive process, at least 32GB RAM recommended.

```{r met-data-se}

# Prepare the methylation SummarizedExperiment from download_dir files
luad_met <- GDCprepare(
  query = luad_query_met,
  directory = download_dir,
)

# Explicitly save as luad_met, otherwise the object is named data
save(luad_met, file = file.path(download_dir, "LUAD_methylation_SE.rda"))

```

## 1.3.1 Option B: LUAD Methylation Data file loading shortcut

Load provided LUAD methylation data SE file to skip download & consolidation steps.

```{r met-data-se-load}

# Specify folder where SE file is
download_dir <- "LUAD_met_data_files"

# Construct full relative path
luad_met_path <- file.path(download_dir, "LUAD_methylation_SE.rda")

# Load the saved SummarizedExperiment object
load(luad_met_path)

```

## 1.3.2 Data Exploration, Preprocessing & Filtering

### **Methylation Probe Quality Control**

Visualise the number of probes flagged by each MASK category, masks are useful for quality control. MASK_general is a comprehensive summary of all other bad-quality flags.

```{r probe-masks-plot}

# Extract all MASK_* columns
mask_cols <- grep("^MASK_", colnames(rowData(luad_met)), value = TRUE)

# Count how many probes are flagged TRUE in each column
mask_counts <- sapply(mask_cols, function(col) {
  sum(rowData(luad_met)[[col]] == TRUE, na.rm = TRUE)
})

# Create data frame for plotting
mask_df <- data.frame(
  Mask_Type = names(mask_counts),
  Num_Probes = as.numeric(mask_counts)
)

# Add a column to mark general mask
mask_df$Highlight <- ifelse(mask_df$Mask_Type == "MASK_general", "General", "Other")

# Clean labels for plot
mask_df$Mask_Label <- gsub("^MASK_", "", mask_df$Mask_Type)
mask_df$Mask_Label <- gsub("_", " ", mask_df$Mask_Label)

# Order for plotting
mask_df <- mask_df[order(mask_df$Num_Probes, decreasing = TRUE), ]
mask_df$Mask_Label <- factor(mask_df$Mask_Label, levels = rev(mask_df$Mask_Label))

# Plot with general mask in red
ggplot(mask_df, aes(x = Mask_Label, y = Num_Probes, fill = Highlight)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("General" = "red", "Other" = "#377eb8")) +
  labs(
    title = "Number of Probes Flagged by Each MASK Category",
    x = "MASK Type",
    y = "Number of Flagged Probes"
    # fill = "Category"  # Removed this line
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

```

#### **Filter out bad quality probes**

```{r bad-probe-filter}

# Filter out MASK_general == TRUE probes, bad quality
luad_met_filtered <- luad_met[!rowData(luad_met)$MASK_general %in% TRUE, ]

# Save luad_met_filtered object for future use

# File path, relative to working dir
path_luad_met_filtered <- file.path("LUAD_met_data_files", "luad_met_filtered.RData")
save(luad_met_filtered, file = path_luad_met_filtered)

```

#### Filtered probe data β values distribution

Distribution of β values after filtering out bad quality probes to verify quality. β values will be transformed in to M-values for downstream analysis.

```{r beta-meta-extract}

# Extract β-matrix and metadata from filtered SE object
beta_mat  <- assay(luad_met_filtered)
meta_meth <- as.data.frame(colData(luad_met_filtered))

```

```{r filtered-beta-value-dist}

# Methylation profile distribution after bad quality probe filtering & pre-sample matching with RNA-seq
plot(density(beta_mat[,1], na.rm = TRUE), 
     col = "grey", main = "Beta Value Distribution", xlab = "Beta", lwd = 2)
for (i in 2:10) {
  lines(density(beta_mat[,i], na.rm = TRUE), col = "grey")
}

```

### Sample matching with RNA-Seq Train dataset

Using TCGA barcodes from methylation data metadata.

```{r sample-match-met-rna}

# Subset to RNA‐seq samples only
rse_train <- rse_luad_sgc_filtered[, colData(rse_luad_sgc_filtered)$split == "train"]
rna_ids <- substr(colData(rse_train)$tcga.cgc_sample_id, 1, 16)
meth_ids <- substr(colnames(beta_mat), 1, 16)
common <- intersect(rna_ids, meth_ids)

# Select only common samples of type Primary_Tumor
primary_tumor_met_samples <- meta_meth$sample[meta_meth$sample_type == "Primary Tumor"]

# Intersect RNA-matched methylation samples with Primary_Tumor samples only
common_primary <- intersect(common, primary_tumor_met_samples)

# Align and subset the β‐matrix with only common primary tumor samples
idx_beta <- match(common_primary, meth_ids)
beta_sub <- beta_mat[, idx_beta]

# Filter RNA-seq rse_train to keep only matching primary tumor samples
rna_ids_primary <- match(common_primary, rse_train$tcga.cgc_sample_id)
rse_train_primary <- rse_train[, rna_ids_primary]

# Save rse_train_primary & common_primary samples 

# File path, relative to working dir
path_train_primary <- file.path("LUAD_rse_objects", "rse_train_primary.RData")
save(rse_train_primary, file = path_train_primary)

path_common_primary <- file.path("LUAD_rse_objects", "common_primary.RData")
save(common_primary, file = path_common_primary)

```

Venn Diagram to visualise methylation and RNA-seq sample intersection.

```{r met-rna-sample-intersection-vd}

# Venn Diagram using rna_ids and meth_ids
vd <- venn.diagram(
    x = list(
        RNA_seq = unique(rna_ids),
        Methylation = unique(meth_ids)
    ),
    filename = NULL,
    fill = c("skyblue", "lightpink"),
    alpha = 0.5,
    cex = 2,
    cat.cex = 1.5,
    main = "Sample Overlap: RNA-seq vs Methylation"
)

grid.newpage()
grid.draw(vd)

```

Visualise intersecting primary tumor samples only.

```{r met-rna-tumor-sample-intersection-vd}

# Intersecting only primary tumor samples

# Subset RNA and methylation IDs to primary tumor samples
rna_primary <- intersect(rna_ids, primary_tumor_met_samples)
meth_primary <- intersect(meth_ids, primary_tumor_met_samples)

# Now plot their Venn diagram
sample_sets_primary <- list(
  RNA_seq_primary = unique(rna_primary),
  Methylation_primary = unique(meth_primary)
)

vd_primary <- venn.diagram(
  x = sample_sets_primary,
  filename = NULL,
  fill = c("skyblue", "lightpink"),
  alpha = 0.5,
  cex = 2,
  cat.cex = 1.5,
  main = "Primary Tumor Sample Overlap: RNA-seq vs Methylation"
)

grid.newpage()
grid.draw(vd_primary)

```

```{r sample-matched-beta-value-dist}

# Shows methylation profile distribution after sample matching & pre-transformation to M-values.
plot(density(beta_sub[,1], na.rm = TRUE), 
     col = "grey", main = "Sample-Matched Beta Value Distribution", xlab = "Beta", lwd = 2)
for (i in 2:10) {
  lines(density(beta_sub[,i], na.rm = TRUE), col = "grey")
}

```

#### β values transformation into M values

```{r beta-mval-transform}

# Convert beta values to M-values for downstream analysis
percentM <- beta_sub * 100
percentM <- pmin(pmax(percentM, 0.01), 99.99)  # clamp to avoid ±Inf
Mval <- log2(percentM / (100 - percentM)) 

```

Visualisation of M-values distribution.

```{r mval-dist}

# M-values distribution (post-transformation)
plot(density(Mval[,1], na.rm = TRUE), 
     col = "grey", main = "M-value Distribution", xlab = "M-value", lwd = 2)
for (i in 2:10) {
  lines(density(Mval[,i], na.rm = TRUE), col = "grey")
}

```

## 1.3.3 Pairwise Comparison Data Subsets Visualisation

The data subsetting for the differetial methylation pairwise comparisons is done in the differential methylation section of this analysis (Data Modeling). As part of data preprocessing, here we visualise the idea behind our subsetting and insight into our sample sizes.

```{r met-pairwise-comp-heatmap}

# Heatmap of methylation data pairwise comparisons

# Get sample counts for subtypes and stages from rse
subtype_counts <- table(rse_train_primary$expression_subtype)
stage_counts <- table(rse_train_primary$stage_collapsed)

# Define pairwise comparisons
pairwise_meth <- tribble(
  ~Comparison,           ~Group1,        ~Group2,
  "proxInflam_vs_proxProlif", "prox.-inflam",   "prox.-prolif.",
  "TRU_vs_proxProlif",        "TRU",          "prox.-prolif.",
  "TRU_vs_proxInflam",        "TRU",          "prox.-inflam",
  "StageI_vs_II",         "Stage_I",      "Stage_II",
  "StageI_vs_III",        "Stage_I",      "Stage_III",
  "StageI_vs_IV",         "Stage_I",      "Stage_IV",
  "StageII_vs_III",       "Stage_II",     "Stage_III",
  "StageII_vs_IV",        "Stage_II",     "Stage_IV",
  "StageIII_vs_IV",       "Stage_III",    "Stage_IV"
) %>%
  # Add sample counts
  rowwise() %>%
  mutate(
    N1 = ifelse(grepl("^Stage", Group1),
                stage_counts[[Group1]],
                subtype_counts[[Group1]]),
    N2 = ifelse(grepl("^Stage", Group2),
                stage_counts[[Group2]],
                subtype_counts[[Group2]])
  ) %>%
  ungroup()

# Melt into long format for plotting
long1 <- melt(pairwise_meth[, c("Comparison", "Group1", "N1")], id.vars = "Comparison")
long2 <- melt(pairwise_meth[, c("Comparison", "Group2", "N2")], id.vars = "Comparison")

long1$Group <- pairwise_meth$Group1
long1$Count <- pairwise_meth$N1

long2$Group <- pairwise_meth$Group2
long2$Count <- pairwise_meth$N2

# Combine both
melted_meth <- rbind(long1[, c("Comparison", "Group", "Count")],
                     long2[, c("Comparison", "Group", "Count")])

# Define display order on heatmap: first subtypes, then stages
comparison_order <- c(
  "proxInflam_vs_proxProlif",
  "TRU_vs_proxProlif",
  "TRU_vs_proxInflam",
  "StageI_vs_II",
  "StageI_vs_III",
  "StageI_vs_IV",
  "StageII_vs_III",
  "StageII_vs_IV",
  "StageIII_vs_IV"
)

# Factor levels for Comparison to enforce order
melted_meth$Comparison <- factor(melted_meth$Comparison, levels = comparison_order)

# Plot
ggplot(melted_meth, aes(x = Comparison, y = Group, fill = Count)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Count), color = "white", size = 3) +
  scale_fill_gradient(low = "orchid", high = "#08306b") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Methylation: Samples per Group in Pairwise Comparisons",
       x = "Comparison", y = "Group") +
  theme(plot.title = element_text(hjust = 0.5))

```

After this first section of the analysis, we have produced preprocessed data that will serve as input for the Data Modeling step of the omics cycle.
