---

title: "LUAD Multi-Omics Classification: Feature Selection and Model Training"
output:
html_document:
df_print: paged
---------------

# Feature Selection and Model Training

## Project Overview

This notebook performs differential expression and differential methylation analyses to identify biologically relevant features associated with LUAD molecular subtypes.

Selected features are integrated into machine learning workflows for subtype classification.

## Pipeline Position

Step 2 of 4

Data Acquisition → Data Integration → Feature Selection & Modeling → Evaluation → Biological Interpretation

## Methods

* Differential Expression Analysis (DESeq2)
* Differential Methylation Analysis (limma)
* Multi-omics feature integration
* Random Forest classification
* Support Vector Machine classification

## Outputs

* Significant DEGs
* Significant DMPs
* Integrated feature matrices
* Trained machine learning models

## Authors

Developed as part of a collaborative academic project at Freie Universität Berlin.


# Section 2: Data Modeling

Our final project focused on working with the TCGA LUAD (Lung Adenocarcinoma) RNA-Seq, Methylation and Clinical data. This R Notebook contains the code corresponding to the Data Modeling, step 3 of the omics data analysis cycle, including the following sections:

3.  **Data Modeling**

    a\. Differential Gene Expression Analysis - Pairwise Comparisons

    b\. Differential Methylation Analysis - Pairwise Comparisons

    c\. Feature Selection - Intersecting genes from DEGs and DMP Results

    d\. Feature Matrices Generation for Machine Learning Classification - Subtype & Stage

    e\. Machine Learning Model Training, Tuning & Evaluation - RF & SVM

# 2.1 Setup

Load required libraries.

```{r setup, message=FALSE, warning=FALSE}

# Global chunk options
knitr::opts_chunk$set(echo = TRUE)

# Install Bioconductor only if not already installed
#if (!requireNamespace("BiocManager", quietly = TRUE))
    #install.packages("BiocManager")

# Bioconductor Packages
#BiocManager::install(c("recount3", "SummarizedExperiment", "DESeq2", "TCGAutils", "TCGAbiolinks", "curatedTCGAData", "sesame", "sesameData", "org.Hs.eg.db"))

# CRAN Packages
#install.packages("tidyverse", "caret", "dplyr", "pheatmap", "tibble", "ggplot2", "reshape2", "pROC")

# Load libraries
library(recount3) #RNA-Seq data
library(SummarizedExperiment) #RNA-seq data manipulation
library(tidyverse) # Data Manipulation
library(DESeq2) #DEG
library(caret) # Stratification
library(dplyr) # Stratification
library(curatedTCGAData) #Cancer subtypes and stage
library(TCGAbiolinks) # Methylation data
library(sesame) #Methylation
library(sesameData) #Methylation
library(org.Hs.eg.db) #Map to HGNC gene symbols
library(randomForest)
library(pROC)
library(e1071)
library(readr)

# Visualization
library(pheatmap)
library(tibble)
library(ggplot2)
library(reshape2)
library(VennDiagram)
library(grid)

```

Set working directory. Intermediate objects will be saved in and loaded from folders created in this working location.

```{r work-dir-setup}

work_dir <- "/home/rstudio/data/FinalProject/LUAD Project RData/Final_Code/"

# Set working directory
setwd(work_dir)

```

# 2.2 Differential Gene Expression Analysis

## 2.2.1 Set up input data and parameters

*DESeq2* Differential Gene Expression analysis will be run in a pairwise manner.

```{r input-param, message=FALSE, warning=FALSE}

# Note: path/file names shoud be change based on comparison pair

rdata_path <- "rse_prad_2etv1_vs_5spop.RData"     # path to the .RData file containing the SummarizedExperiment object
rse_object_name <- "rse_prad_2etv1_vs_5spop"      # name of the object to load from the RData file

group_var <- "expression_subtype"                # column in colData used to define experimental groups

condition_1 <- "2-ETV1"                          # first condition (will be used as numerator in log2FC)
condition_2 <- "5-SPOP"                          # second condition (will be used as denominator in log2FC)

```

## 2.2.2 Set up output directories

```{r out-dir-setup, message=FALSE, warning=FALSE}

#Output base directory
out_dir <- "/Users/mac/Desktop/data_science_project"

# Preparing folder and file naming
file_base <- tools::file_path_sans_ext(basename(rdata_path))
prefix <- sub("^rse_", "", file_base)
comparison_label <- paste0(condition_1, "_vs_", condition_2)
comparison_label_safe <- gsub("[^A-Za-z0-9_\\-]", "_", comparison_label)

# Creating sub-directory for saving results
sub_dir <- file.path(out_dir, prefix)
if (!dir.exists(sub_dir)) dir.create(sub_dir, recursive = TRUE)

```

## 2.2.3 Differential Expression Analysis

```{r dseq2-analysis, message=FALSE, warning=FALSE}

#Load the RData object
load(rdata_path)

# Ensure the object exists
if (!(rse_object_name %in% ls())) {
  stop(paste("Object", rse_object_name, "not found in loaded environment"))
}

rse <- base::get(rse_object_name)

# Filter samples with non-NA group labels
coldata <- colData(rse)
valid_samples <- !is.na(coldata[[group_var]])
rse_filtered <- rse[, valid_samples]

# Run DESeq2 analysis
dds <- DESeqDataSetFromMatrix(
  countData = assays(rse_filtered)$raw_counts,
  colData = colData(rse_filtered),
  design = as.formula(paste("~", group_var))
)
dds <- DESeq(dds)

# Get results for the desired contrast
res <- results(dds, contrast = c(group_var, condition_1, condition_2), alpha = 0.05)
res_df <- as.data.frame(res)
res_df <- res_df[complete.cases(res_df[, c("padj", "log2FoldChange")]), ]

# Filter significant DEGs
# We define significant differentially expressed genes (DEGs) using the following thresholds:
# - Adjusted p-value (padj) < 0.05: controls for multiple testing (FDR < 5%)
# - Absolute log2 fold change (|log2FC|) ≥ 1: corresponds to at least 2-fold up/down regulation

deg_df <- res_df[which(res_df$padj < 0.05 & abs(res_df$log2FoldChange) >= 1), ]
deg_df$gene <- rownames(deg_df) # Add gene names as a separate column (they are currently row names)
deg_df <- deg_df[order(deg_df$padj), ] # Sort the DEGs by adjusted p-value in ascending order (most significant genes first)

# Normalize counts
# Getting normalized counts using DESeq2's size factor normalization
# These values are useful for visualization

norm_counts <- counts(dds, normalized = TRUE)

# Apply variance stabilizing transformation (VST) to the count data
vsd <- varianceStabilizingTransformation(dds, blind = TRUE)
vsd_mat <- assay(vsd)

# Extract sample metadata as a plain data frame (for annotation)
metadata_df <- as.data.frame(colData(dds))

# Save outputs
saveRDS(dds, file = file.path(sub_dir, paste0(prefix, "_", comparison_label_safe, "_dds_processed.rds")))
saveRDS(res, file = file.path(sub_dir, paste0(prefix, "_", comparison_label_safe, "_deseq2_results_full.rds")))
saveRDS(vsd, file = file.path(sub_dir, paste0(prefix, "_", comparison_label_safe, "_vsd_object.rds")))

write.csv(res_df, file = file.path(sub_dir, paste0(prefix, "_", comparison_label_safe, "_deseq2_results_full.csv")), row.names = TRUE)
write.csv(deg_df, file = file.path(sub_dir, paste0(prefix, "_", comparison_label_safe, "_DEG_filtered.csv")), row.names = FALSE)
write.csv(norm_counts, file = file.path(sub_dir, paste0(prefix, "_", comparison_label_safe, "_normalized_counts.csv")))
write.csv(vsd_mat, file = file.path(sub_dir, paste0(prefix, "_", comparison_label_safe, "_vsd_log_normalized_counts.csv")))
write.csv(metadata_df, file = file.path(sub_dir, paste0(prefix, "_", comparison_label_safe, "_sample_metadata.csv")))

```

## 2.2.4 Results Visualisation

MA, Volcano and Heatmap plots.

```{r visualization, message=FALSE, warning=FALSE}

# MA Plot
png(filename = file.path(sub_dir, paste0(prefix, "_", comparison_label_safe, "_MA_plot.png")), width = 900, height = 600, res = 120)
plotMA(res, ylim = c(-5, 5), main = paste(prefix, comparison_label, ":", "MA Plot"))
dev.off()

# Volcano Plot
res_df$threshold <- as.factor(abs(res_df$log2FoldChange) > 1 & res_df$padj < 0.05)
volcano_plot <- ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = threshold)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("grey", "red")) +
  theme_minimal() +
  labs(x = "log2 Fold Change", y = "-log10(adjusted p-value)",
       title = paste(prefix, comparison_label, ":", "Volcano Plot")) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5))
ggsave(filename = file.path(sub_dir, paste0(prefix, "_", comparison_label_safe, "_Volcano_plot.png")),
       plot = volcano_plot, width = 7, height = 6, dpi = 120)

# Heatmap of Top 30 DEGs
top_genes <- head(rownames(deg_df), 30)
heatmap_data <- norm_counts[top_genes, ]
heatmap_data <- t(scale(t(heatmap_data)))

annotation_col <- metadata_df[, group_var, drop = FALSE]
colnames(annotation_col) <- "Group"

png(filename = file.path(sub_dir, paste0(prefix, "_", comparison_label_safe, "_Heatmap_top30_genes.png")), width = 1000, height = 700, res = 120)
pheatmap(heatmap_data,
         annotation_col = annotation_col,
         show_rownames = TRUE,
         show_colnames = FALSE,
         clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation",
         fontsize_row = 8,
         main = paste(prefix, comparison_label, ":", "Top 30 DEGs Heatmap"))
dev.off()

```

# 2.3 Differential Methylation Analysis

Pairwise differential methylation analysis between stages and subtypes. \
**Note:** some pre-processing steps are rerun to produce contrast matrices for *limma*.

```{r met-data-load}

# Load the saved object (so future knits skip the heavy prepare)
load(file.path(download_dir, "LUAD_methylation_SE.rda"))  # gives `luad_met`

# Filter out MASK_general == TRUE probes, as these are bad quality
luad_met_filtered <- luad_met[!rowData(luad_met)$MASK_general %in% TRUE, ]

```

```{r met-data-processing}

# Extract β-matrix and metadata
beta_mat  <- assay(luad_met_filtered)
meta_meth <- as.data.frame(colData(luad_met_filtered))

# Subset to RNA‐seq samples only
rse_train <- rse_luad_sgc_filtered[, colData(rse_luad_sgc_filtered)$split == "train"] #493 train samples from RNA-seq
rna_ids   <- substr(colData(rse_train)$tcga.cgc_sample_id, 1, 16)
meth_ids  <- substr(colnames(beta_mat), 1, 16)
common <- intersect(rna_ids, meth_ids) # 392 common samples between meth and RNA-seq

# Select only common samples of type primary tumor
primary_tumor_met_samples <- meta_meth$sample[meta_meth$sample_type == "Primary Tumor"] 

# Intersect RNA-matched methylation samples with primary tumors only
common_primary <- intersect(common, primary_tumor_met_samples) 

# Align and subset the β‐matrix with only common primary tumor samples
idx_beta <- match(common_primary, meth_ids)
beta_sub <- beta_mat[, idx_beta]

# Filter RNA-seq rse_train to keep only matching primary tumor samples
rna_ids_primary <- match(common_primary, rse_train$tcga.cgc_sample_id)
rse_train_primary <- rse_train[, rna_ids_primary]

# Convert beta values to M-values once for both stage and subtype comparisons
percentM <- beta_sub * 100
percentM <- pmin(pmax(percentM, 0.01), 99.99)  # clamp to avoid ±Inf
Mval <- log2(percentM / (100 - percentM)) 

# Build the stage factor in the same order
stages_train <- factor(
  rse_train_primary$stage_collapsed[match(common_primary, rse_train_primary$tcga.cgc_sample_id)],
  levels = c("Stage_I", "Stage_II", "Stage_III", "Stage_IV")
)

# Subtype factor in the same order
subtypes_train <- factor(
  rse_train_primary$expression_subtype[match(common_primary, rse_train_primary$tcga.cgc_sample_id)],
  levels = c("prox.-inflam", "prox.-prolif.", "TRU"),
  labels = c("prox_inflam", "prox_prolif", "TRU") # for readability
)

# Subtype NA filtering
keep_subtype <- !is.na(subtypes_train)
Mval_subtype <- Mval[, keep_subtype]
subtypes_train_filtered <- subtypes_train[keep_subtype]

# Stage NA filtering
keep_stage <- !is.na(stages_train)
Mval_stage <- Mval[, keep_stage]
stages_train_filtered <- stages_train[keep_stage]

```

*limma* contrast matrices creation to specify pairwise comparisons for subtype and stage.

```{r subtype-limma-contrast-matrix, message=FALSE}

# Set up limma design & contrasts

# Subtype
# Design matrix, no intercept
design_subtype <- model.matrix(~ 0 + subtypes_train_filtered)
colnames(design_subtype) <- levels(subtypes_train_filtered)

# Pairwise subtype comparisons, using only the filtered primary tumor samples
contMat_subtype <- makeContrasts(
  proxProlif_vs_proxInflam = prox_prolif - prox_inflam,
  TRU_vs_proxInflam        = TRU         - prox_inflam,
  TRU_vs_proxProlif        = TRU         - prox_prolif,
  levels = design_subtype
)

# Fit model
fit_subtype <- lmFit(Mval_subtype, design_subtype)
fit_subtype <- contrasts.fit(fit_subtype, contMat_subtype)
fit_subtype <- eBayes(fit_subtype) 

```

```{r stage-limma-contrast-matrix, message=FALSE}

# Stage

# Design matrix, no intercept
design_stage <- model.matrix(~ 0 + stages_train_filtered)
colnames(design_stage) <- levels(stages_train_filtered)

# Subtype pairwise comparisons
contMat_stage <- makeContrasts(
  II_vs_I   = Stage_II   - Stage_I,
  III_vs_I  = Stage_III  - Stage_I,
  IV_vs_I   = Stage_IV   - Stage_I,
  III_vs_II = Stage_III  - Stage_II,
  IV_vs_II  = Stage_IV   - Stage_II,
  IV_vs_III = Stage_IV   - Stage_III,
  levels = design_stage
)

# Fit model
fit_stage <- lmFit(Mval_stage, design_stage)
fit_stage <- contrasts.fit(fit_stage, contMat_stage)
fit_stage <- eBayes(fit_stage)

```

Extract differentially methylated probes (DMPs) from each comparison.

```{r dm-extraction}

# Extract DMPs for each contrast and save

# Create gene annotation vector using luad_met metadata
gene_annot_vector <- rowData(luad_met_filtered)$gene_HGNC #421433, corresponds to genes mapped to probes
names(gene_annot_vector) <- rownames(luad_met_filtered)

# Helper function to extract and save dmp results
save_dmp_results <- function(fit_obj, contMat, outdir_prefix, gene_annot_vector) {
  if (!dir.exists(outdir_prefix)) dir.create(outdir_prefix)
  
  for (coef in colnames(contMat)) {
    dmp <- topTable(fit_obj, coef = coef, number = Inf, adjust.method = "fdr")
    
    # Add gene_HGNC annotation
    dmp$gene_HGNC <- gene_annot_vector[rownames(dmp)]
    
    # Split probes that map to multiple genes
    dmp_expanded <- dmp |>
      tibble::rownames_to_column("probe_id") |>
      tidyr::separate_rows(gene_HGNC, sep = ";") |>
      dplyr::filter(!is.na(gene_HGNC) & gene_HGNC != "")
    
    # Save as .csv and .rds
    write.csv(dmp_expanded,
              file = file.path(outdir_prefix, paste0("DMP_", coef, ".csv")),
              row.names = FALSE)
    
    saveRDS(dmp_expanded,
            file = file.path(outdir_prefix, paste0("DMP_", coef, ".rds")))
  }
}

# Using the function to produce and save subtype and stage results
save_dmp_results(fit_subtype, contMat_subtype, "LUAD_methylation_limma_subtype_results", gene_annot_vector)
save_dmp_results(fit_stage, contMat_stage, "LUAD_methylation_limma_stage_results", gene_annot_vector) 

```

# 2.4 Feature Selection

Intersection of differentially methylated probes and differentially expressed genes resulting from *DESeq2* and *limma* runs.

```{r gene-intersection-prep}

# Ensembl to symbol mapping function
ensembl_to_symbol <- function(ensembl_ids) {
  ids_noversion <- sub("\\..*$", "", ensembl_ids)
  map <- AnnotationDbi::select(org.Hs.eg.db,
                               keys = ids_noversion,
                               keytype = "ENSEMBL",
                               columns = c("SYMBOL"))
  map <- map[!is.na(map$SYMBOL), ]
  symbol_vec <- map$SYMBOL
  names(symbol_vec) <- map$ENSEMBL
  unique(symbol_vec[ids_noversion])
}

# Location of DEGs and DMPs
de_dir <- "Differential expression"
meth_subtype_dir <- "Differential Methylation Cluster Results/LUAD_methylation_limma_subtype_results"
meth_stage_dir   <- "Differential Methylation Cluster Results/LUAD_methylation_limma_stage_results"
outdir <- "LUAD_intersections"
if (!dir.exists(outdir)) dir.create(outdir)

```

```{r deg-dmp-files-processing}

# Find all DEG and DMP .csv files 
de_files <- list.files(de_dir, pattern = "\\.csv$", full.names = TRUE)
meth_files <- c(
  list.files(meth_stage_dir, pattern = "\\.csv$", full.names = TRUE),
  list.files(meth_subtype_dir, pattern = "\\.csv$", full.names = TRUE)
)

# Helper function: extract comparison name
get_comp <- function(path) {
  name <- tools::file_path_sans_ext(basename(path))
  sub("^DMP_", "", name)  # Remove "DMP_" from start if present to match DE files
}

```

Intersect significantly expressed and methylated genes.

```{r get-sig-results}
get_sig_degs <- function(de_path, fdr_col = "padj", gene_col = "gene", threshold = 0.05) {
  dat <- read_csv(de_path, show_col_types = FALSE)
  dat %>% filter(!is.na(!!sym(fdr_col)), !!sym(fdr_col) < threshold) %>%
    pull(!!sym(gene_col)) %>% unique()
}

get_sig_dmps <- function(meth_path, fdr_col = "adj.P.Val", gene_col = "gene_HGNC", threshold = 0.05) {
  dat <- read_csv(meth_path, show_col_types = FALSE)
  # flatten out comma-separated symbols 
  sig_genes <- dat %>%
    filter(!is.na(!!sym(fdr_col)), !!sym(fdr_col) < threshold,
           !is.na(!!sym(gene_col)), !!sym(gene_col) != "") %>%
    pull(!!sym(gene_col)) %>%
    str_split(",") %>% unlist() %>% str_trim() %>% unique()
  sig_genes[sig_genes != ""]
}
```

Loop over all comparisons: produce intersection files.

```{r intersection-file-generation}

# Create tibbles for easy joining by comp name
de_tbl <- tibble(de_path = de_files, comp = get_comp(de_files))
meth_tbl <- tibble(meth_path = meth_files, comp = get_comp(meth_files))

# Define which comparison names are stages
stage_names <- c("II_vs_I", "III_vs_I", "IV_vs_I", "III_vs_II", "IV_vs_II", "IV_vs_III")

# Initialize lists for union tracking
intersect_union_subtypes <- list()
intersect_union_stages <- list()

for(i in seq_len(nrow(de_tbl))) {
  comp <- de_tbl$comp[i]
  de_path <- de_tbl$de_path[i]
  
  # Try to find meth file with exactly the same comp name
  meth_match <- filter(meth_tbl, comp == !!comp)
  if(nrow(meth_match) == 0) {
    cat("No methylation CSV for:", comp, "<br>\n")
    next
  }
  meth_path <- meth_match$meth_path[1]

  degs <- get_sig_degs(de_path)
  degs_symbols <- toupper(ensembl_to_symbol(degs))
  dmps <- toupper(get_sig_dmps(meth_path))
  intersection <- intersect(degs_symbols, dmps)
  
  # Determine if this is a stage or subtype comparison for union
  if (comp %in% stage_names) {
    intersect_union_stages[[comp]] <- intersection
  } else {
    intersect_union_subtypes[[comp]] <- intersection
  }

  # Output summary
  cat("###", comp, "<br>\n")
  cat("- Significant DEGs:", length(degs), "<br>\n")
  cat("- Significant DMPs:", length(dmps), "<br>\n")
  cat("- Genes in both:", length(intersection), "<br>\n")
  if(length(intersection) > 0) print(knitr::kable(head(intersection, 10), col.names = "Intersected Genes"))

  # Venn Diagrams
  grid.newpage()
  draw.pairwise.venn(
    area1 = length(degs), area2 = length(dmps), cross.area = length(intersection),
    category = c("DEGs", "DMPs"), fill = c("red", "blue"),
    main = paste("Overlap in", comp)
  )
  # Save
  safe_comp <- gsub("[^a-zA-Z0-9]", "_", comp)
  write_csv(data.frame(Gene = intersection),
            file = file.path(outdir, paste0("Intersection_", safe_comp, ".csv")))
}

# Save union of intersected genes for multiclass classifiers
union_subtype_genes <- unique(unlist(intersect_union_subtypes))
union_stage_genes   <- unique(unlist(intersect_union_stages))

write_csv(tibble(Gene = union_subtype_genes), file = file.path(outdir, "Union_Subtype_Intersected_Genes.csv"))
write_csv(tibble(Gene = union_stage_genes),   file = file.path(outdir, "Union_Stage_Intersected_Genes.csv"))

```

# 2.5 Feature Matrices Generation

For feature matrix generation, first a global matrix integrating methylation and expression values will be created and scaled. Then, we will create feature matrices to serve as input for our binary classification and multi-class classifiers, using features found in the feature selection step for each pairwise comparison. After intersection, only subtype comparisons were kept in the analysis, as no DEG and DMP intersecting genes were found for the stage comparisons.

## 2.5.1 Data Normalization & Processing

Differential gene expression was done on raw counts, methylation data was transformed into M values. For model training, expression counts will be normalized with *DESEq2::vst* function, and M values will be used for methylation data.

### RNA-Seq Data Processing

```{r rna-fm-object-load}

# Load objects from previous processing steps that are needed for feature matrix generation

# rse_train_primary, contains primary tumor training samples
# common_primary, list of common samples between RNA-sequences & methylation data
# Specify folder where RSE file is
rse_dir <- "LUAD_rse_objects"

# Construct full relative path
rse_path <- file.path(rse_dir, "rse_train_primary.RData")
common_primary_path <- file.path(rse_dir, "common_primary.RData")

# Load the saved object
load(rse_path)
load(common_primary_path)

```

Create a normalized data expression matrix with samples as column names.

```{r rna-data-norm, message=FALSE}

# Build subtype factors from rse_train_primary (if not available from differential methylation)
subtypes_train <- factor(
  rse_train_primary$expression_subtype[match(common_primary, rse_train_primary$tcga.cgc_sample_id)],
  levels = c("prox.-inflam", "prox.-prolif.", "TRU"),
  labels = c("prox_inflam", "prox_prolif", "TRU") # for readability
)

# Subtype NA filtering
keep_subtype <- !is.na(subtypes_train)
Mval_subtype <- Mval[, keep_subtype] 
subtypes_train_filtered <- subtypes_train[keep_subtype]

# Subset rse_train_primary to only contain samples with subtype
short_ids <- stringr::str_sub(rse_train_primary$tcga.cgc_sample_id, 1, 12)
rse_train_primary$short_id <- short_ids
keep_samples <- short_ids %in% names(subtypes_train_filtered)
rse_train_subtyped <- rse_train_primary[, keep_samples]

# Expression counts: Normalization using DESeq2::vst
raw_counts <- assays(rse_train_subtyped)$raw_counts

vst_obj <- vst(DESeqDataSetFromMatrix(
  countData = raw_counts,
  colData   = colData(rse_train_subtyped),
  design    = ~ 1
))

```

Create RNA-Seq normalized expression data matrix.

```{r rna-exp-matrix}

# Create normalized counts expression matrix, rows = genes, cols = samples
expr_mat <- assay(vst_obj)  # vst-normalized expression

# Assign TCGA sample ids as colnames for expr_mat
colnames(expr_mat) <- rse_train_subtyped$tcga.cgc_sample_id

```

### Ensemble IDs to HGNC Gene symbol mapping

RNA-Seq has ENSEMBLE gene IDs, they will be mapped to HGNC gene symbols.\
Multiple ENSEMBL IDs map to a single HGNC gene symbol, for our feature matrix, a single expression value will be present per gene, so records mapping to the same gene will be averaged to get a single expression value per gene.

```{r ensembl-to-symbol}

# ensembl_to_symbol function: convert RNA-seq gene Ensemble ids to HGNC,returns a named vector with mappings, (not unique genes as done for intersection)

ensembl_to_symbol <- function(ensembl_ids) {
  ids_noversion <- sub("\\..*$", "", ensembl_ids)
  
  map <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys    = unique(ids_noversion),
    keytype = "ENSEMBL",
    columns = "SYMBOL"
  )
  
  map <- map[!is.na(map$SYMBOL) & map$SYMBOL != "", ]
  map <- map[!duplicated(map$ENSEMBL), ]

  # Named vector: names = Ensembl, values = SYMBOL
  hgnc_vec <- setNames(map$SYMBOL, map$ENSEMBL)

  # Return in input order
  return(hgnc_vec[ids_noversion])
}

```

```{r rna-gene-symbol-map, message=FALSE}

# Remove version from ID
ensembl_ids <- gsub("\\..*", "", rownames(expr_mat))  # Remove version
hgnc_names <- ensembl_to_symbol(ensembl_ids)
gene_symbols <- hgnc_names[ensembl_ids]

# Add gene symbols as a column to a df before assigning as rownames
expr_df <- as.data.frame(expr_mat)
expr_df$gene_symbol <- gene_symbols

# Move gene_symbol to the first column
expr_df <- expr_df[, c(ncol(expr_df), 1:(ncol(expr_df) - 1))]

#Convert to tibble
expr_df <- as_tibble(expr_df)

# Filter NA gene symbols and aggregate duplicated gene expression values
expr_df <- expr_df %>%
  filter(!is.na(gene_symbol) & gene_symbol != "")

# Average duplicates
expr_avg <- expr_df %>%
  group_by(gene_symbol) %>%
  summarise(across(everything(), mean), .groups = "drop")

# Replace expr_mat with expr_avg to use downstream as matrix again
expr_mat <- expr_avg %>%
  column_to_rownames("gene_symbol") %>%
  as.matrix()

```

### Methylation Data Processing

```{r met-fm-object-load}

# Load objects from previous processing steps needed for feature matrix generation

# Specify folder where luad_met_filtered file is
luad_met_filtered_dir <- "LUAD_met_data_files"

# Construct full relative path
luad_met_filtered_path <- file.path(luad_met_filtered_dir, "luad_met_filtered.RData")

# Load the saved SummarizedExperiment object
load(luad_met_filtered_path)

```

```{r met-fm-object-creation}

# Create gene annotation vector using luad_met metadata
gene_annot_vector <- rowData(luad_met_filtered)$gene_HGNC
names(gene_annot_vector) <- rownames(luad_met_filtered)

# Recode of subtype names in subtypes_train_filtered
subtypes_train_filtered <- recode(subtypes_train_filtered,
  "prox_prolif" = "proxProlif",
  "prox_inflam" = "proxInflam"
)

```

Collapsing multiple CpG probes that map to one gene, one methylation value per gene will be kept.

```{r met-annot-tbl}

# annot_tbl, one probe per row mapped to corresponding gene
# probes mapping to more than one gene get their own row
annot_tbl <- tibble(
  probe = names(gene_annot_vector),
  gene  = gene_annot_vector
) %>%
  separate_rows(gene, sep = ";") %>%
  filter(!is.na(gene) & gene != "") %>%
  filter(probe %in% rownames(Mval_subtype))

```

```{r met-probe-collapse}

# mval_long, get gene level M values
mval_long <- Mval_subtype[annot_tbl$probe, , drop = FALSE] %>%
  as.data.frame() %>%
  rownames_to_column("probe") %>%
  left_join(annot_tbl, by = "probe") %>%
  filter(!is.na(gene) & gene != "") %>%
  pivot_longer(-c(probe, gene), names_to = "sample", values_to = "mval") %>%
  group_by(gene, sample) %>%
  summarise(meth = mean(mval, na.rm = TRUE), .groups = "drop")

```

Methylation M values matrix creation.

```{r met-mval-matrix}

# Methylation values matrix, rows = genes, columns = samples
meth_mat <- mval_long %>%
  pivot_wider(names_from = sample, values_from = meth) %>%
  column_to_rownames("gene") %>%
  as.matrix()

# Shortening sample names to match expression matrix
colnames(meth_mat) <- substr(colnames(meth_mat), 1, 16)

```

### Gene Intersection between expr_math and meth_mat

Create a set of genes present in both methylation and RNA-seq data, necessary to generate a combined expression + methylation feature matrix.

```{r exp-met-mat-gene-intersect}

# Intersect genes present in both omics, to create a joint matrix
common_genes <- intersect(rownames(expr_mat), rownames(meth_mat)) 

expr_mat_sub <- expr_mat[common_genes, ]
meth_mat_sub <- meth_mat[common_genes, ]

expr_mat <- expr_mat_sub
meth_mat <- meth_mat_sub

cat("Common genes between methylation and expression data:", length(common_genes))
```

### Global Feature Matrix Scaling

The expression and methylation matrices will be combined to get a multi-omics matrix, which will be scaled. This scaled combined matrix will serve as the starting point to subset the features selected during the Feature selection step to prepare the input for each binary and multi-class classifiers.

```{r feature-matrix-scaling}

# Matrix transposing, scaling works when features are cols
expr_mat <- t(expr_mat)
meth_mat <- t(meth_mat)

# Keep omics type in colnames to identify features
colnames(expr_mat) <- paste0("expr_", colnames(expr_mat))
colnames(meth_mat) <- paste0("meth_", colnames(meth_mat))

# Combine both matrix and standardize omics values (scale)
combined_mat <- cbind(expr_mat, meth_mat)

# Scale matrix
scaled_combined_mat <- scale(combined_mat)

cat("The combined scaled feature matrix contains", 
    nrow(scaled_combined_mat), "samples and", 
    ncol(scaled_combined_mat), "features.\n")

```

```{r save-train-mean-sd-mat}

# Create "feature_matrices" folder if it does not exist
feature_mat_dir <- "LUAD_feature_matrices"
if (!dir.exists(feature_mat_dir)) dir.create(feature_mat_dir)

# Save scaled combined matrix in feature_mat_dir
saveRDS(scaled_combined_mat, 
        file = file.path(feature_mat_dir, "LUAD_scaled_expr_meth_mat.rds"))

# Extract and save mean and standard deviation to scale test data
train_means <- attr(scaled_combined_mat, "scaled:center")
train_sds <- attr(scaled_combined_mat, "scaled:scale")

# Save to the feature_mat_dir
saveRDS(train_means, file = file.path(feature_mat_dir, "train_scaling_means.rds"))
saveRDS(train_sds, file = file.path(feature_mat_dir, "train_scaling_sds.rds"))

```

### Binary & Multi-class Classifier Feature Matrices Generation

Each of our pairwise comparisons for methylation and gene expression produced a set of features for model training. This will be used to prepare feature matrices for binary classifiers. The union of the gene intersection will be used to generate a feature matrix for a multi-class classifier.

```{r binary-feat-matrix}

# Build final feature matrices from intersections and union gene sets

# Define directories
inter_dir <- "LUAD_intersections"
out_dir <- "LUAD_feature_matrices"
log_file <- "LUAD_feature_matrices/feature_matrix_log.txt"

# Open log connection
log_con <- file(log_file, open = "wt")
sink(log_con, type = "output")

# List gene intersection and union files
inter_files <- list.files(inter_dir, pattern = "\\.csv$", full.names = TRUE)

for (f in inter_files) {
  comp <- tools::file_path_sans_ext(basename(f)) %>%
    gsub("^Intersection_", "", .) %>%
    gsub("^Union_", "union_", .)
  
  cat("\n>>> Processing:", comp, "\n")

  genes <- read_csv(f, show_col_types = FALSE)$Gene %>% toupper()
  
  # Match columns in combined matrix (columns look like expr_X or meth_X)
  expr_cols <- paste0("expr_", genes)
  meth_cols <- paste0("meth_", genes)
  gene_keep <- intersect(colnames(scaled_combined_mat), c(expr_cols, meth_cols))
  
  cat(" - Genes in input file:", length(genes), "\n")
  cat(" - Matched columns in combined matrix:", length(gene_keep), "\n")
  
  if (length(gene_keep) < 5) {
    cat(" !! Too few matched columns for", comp, "\n")
    next
  }
  
  # Subset feature matrix by genes
  sub_df <- scaled_combined_mat[, gene_keep, drop = FALSE]
  sub_df <- as.data.frame(sub_df)
  
  # Match sample short IDs to subtype mapping
  sub_df$sample <- rownames(sub_df)
  sub_df$short_id <- substr(sub_df$sample, 1, 12)
  sub_df$subtype <- subtypes_train_filtered[sub_df$short_id]
  
  # Filter to keep only labeled samples
  sub_df <- sub_df[!is.na(sub_df$subtype), ]
  
  # For intersections: filter to the two subtypes for pairwise comparison
  if (grepl("^Intersection_", basename(f))) {
    subtype_pair <- unlist(strsplit(comp, "_vs_"))
    sub_df <- sub_df[sub_df$subtype %in% subtype_pair, ]

    if (length(unique(sub_df$subtype)) != 2) {
      cat(" !! Warning: expected 2 subtypes, found:", paste(unique(sub_df$subtype), collapse = ", "), "\n")
    }
  }
  
  # Clean up: drop short_id, reorder
  sub_df$short_id <- NULL
  sub_df <- sub_df[, c(setdiff(colnames(sub_df), c("sample", "subtype")), "sample", "subtype")]

  cat(" - Samples:", nrow(sub_df), "\n")
  cat(" - Features (expr + meth):", length(gene_keep), "\n")
  cat(" - Total columns (incl. sample + label):", ncol(sub_df), "\n")

  # Save final matrix
  write_csv(sub_df, file.path(out_dir, paste0("feature_matrix_", comp, ".csv")))
  saveRDS(sub_df, file.path(out_dir, paste0("feature_matrix_", comp, ".rds")))
  cat(" - Saved to:", file.path(out_dir, paste0("feature_matrix_", comp, ".csv")), "\n")
}

# Close log
sink()
close(log_con)

```

# 2.6 Machine Learning

## 2.6.1 Random Forest - Training, Tuning, Evaluation

We trained a binary classifier per pairwise comparison, initially generating three binary Random Forest Classifiers.

### proxProlif vs proxInflam Binary Classifier Training

This classifier discriminates between subtype classes proxProlif and proxInflam based on methylation and expression data.

#### Load Feature Matrix and Labels

Feature matrix loading.

```{r load-fm-pp-pi}

set.seed(123)

# Set the path to each LUAD training feature matrix
feature_matrix_path <- "Feature Matrices Real Data/feature_matrix_proxProlif_vs_proxInflam.csv"
label_col <- "subtype"  

data <- read_csv(feature_matrix_path)
str(data)
head(data)

# Remove sample column if present
if ("sample" %in% colnames(data)) {
  data <- data %>% dplyr::select(-sample)
}
# Remove rows where subtype is NA
data <- data %>% filter(!is.na(!!sym(label_col)))
data[[label_col]] <- as.factor(data[[label_col]])

```

Model training.

```{r rf-train-pp-pi}

# Remove columns that are all NA
data <- data[, colSums(is.na(data)) < nrow(data)]

# Make sure all features are numeric except label
num_cols <- sapply(data, is.numeric)
num_cols[label_col] <- FALSE
data <- data[, num_cols | names(data) == label_col]

# Drop columns (genes) with >20% NAs, then impute the rest
na_threshold <- 0.2 * nrow(data)
data_filtered <- data[, colSums(is.na(data)) < na_threshold]

# Impute remaining NAs with median
data_imputed <- data_filtered %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm=TRUE), .)))

# Train Random Forest on all imputed data

# Weighted model to deal with moderate class imbalance
rf_model_w <- ranger(
  x = data_imputed %>% dplyr::select(-all_of(label_col)),
  y = data_imputed[[label_col]],
  num.trees = 500,
  importance = "permutation",
  class.weights = c(proxInflam = 1, proxProlif = 1.5)
)

rf_pred_obj <- predict(rf_model_w, data_imputed %>% dplyr::select(-all_of(label_col)))

# Extract predictions, named list
rf_pred_w <- rf_pred_obj$predictions

# Make sure predicted and true labels are factors with same levels
rf_pred_w <- factor(rf_pred_w, levels = levels(data_imputed[[label_col]]))
true_labels <- factor(data_imputed[[label_col]], levels = levels(data_imputed[[label_col]]))

# Confusion Matrix
confusionMatrix(rf_pred_w, true_labels)

```

Cross-validation, tune model parameters and save final model.

```{r rf-cv-pp-pi}

cv_folds <- trainControl(method = "cv", number = 5)

# Define mtry values to search over
tune_grid <- expand.grid(
  mtry = seq(30, 90, by = 15),
  splitrule = "gini",
  min.node.size = 5
)

# Cross-validation with grid search using ranger with class weights
rf_cv <- train(
  x = data_imputed %>% dplyr::select(-all_of(label_col)),
  y = data_imputed[[label_col]],
  method = "ranger",
  trControl = cv_folds,
  tuneGrid = tune_grid,
  num.trees = 500,
  importance = "permutation", 
  weights = ifelse(data_imputed[[label_col]] == "proxProlif", 1.5, 1)
)

print(rf_cv)
plot(varImp(rf_cv))

# Retrain model with best mtry found by cv
best_mtry <- rf_cv$bestTune$mtry
best_splitrule <- rf_cv$bestTune$splitrule
best_min_node_size <- rf_cv$bestTune$min.node.size

rf_model_final <- ranger(
  x = data_imputed %>% dplyr::select(-all_of(label_col)),
  y = data_imputed[[label_col]],
  num.trees = 500,
  mtry = best_mtry,
  splitrule = best_splitrule,
  min.node.size = best_min_node_size,
  importance = "permutation",
  classification = TRUE,
  class.weights = c(proxInflam = 1, proxProlif = 1.5)
)

# Predict on training data, sanity check
rf_pred_obj_final <- predict(rf_model_final, data_imputed %>% dplyr::select(-all_of(label_col)))

# Extract predictions, named list
rf_pred_final <- rf_pred_obj_final$predictions

# Make sure predicted and true labels are factors with same levels
rf_pred_final <- factor(rf_pred_final, levels = levels(data_imputed[[label_col]]))
true_labels <- factor(data_imputed[[label_col]], levels = levels(data_imputed[[label_col]]))

# Confusion Matrix
confusionMatrix(rf_pred_final, true_labels)

saveRDS(rf_model_final, file = "LUAD_Random_Forests/rf_model_final_proxProlif_vs_proxInflam.rds")
```

#### Feature Importance

Extract most important features for model accuracy, input for annotation step.

```{r rf-pp-pi-feat_imp}

# Extract feature importance from the ranger rf model final
importance_tbl <- rf_model_final$variable.importance

# Convert to data frame, sort descending and scale importance addition
importance_df <- data.frame(
  feature = names(importance_tbl),
  importance = as.numeric(importance_tbl)
) %>%
  arrange(desc(importance)) %>%
  mutate(
    scaled_importance = (importance - min(importance)) / (max(importance) - min(importance))
  )

# Save results
write.csv(importance_df, file = "important_features_proxProlif_vs_proxInflam.csv", row.names = FALSE)
saveRDS(importance_df, file = "important_features_proxProlif_vs_proxInflam.rds")

```

```{r rf-pp-pi-feat-imp-plot}
# Feature Importance Plot
library(ggplot2)
top_n <- 20
ggplot(head(importance_df, top_n), aes(x = reorder(feature, importance_df[[1]]), y = importance_df[[1]])) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top Variable Importances", x = "Feature", y = "Importance (Permutation)")
```

### TRU vs proxInflam Binary Classifier Training

This classifier discriminates between subtype classes TRU and proxInflam based on methylation and expression data.

#### Load Feature Matrix and Labels

```{r load-fm-tru-pi}

set.seed(123)

# Set the path to each LUAD training feature matrix
feature_matrix_path <- "Feature Matrices Real Data/feature_matrix_TRU_vs_proxInflam.csv"
label_col <- "subtype"  

data <- read_csv(feature_matrix_path)
str(data)
head(data)

# Remove sample column if present
if ("sample" %in% colnames(data)) {
  data <- data %>% dplyr::select(-sample)
}
# Remove rows where subtype is NA
data <- data %>% filter(!is.na(!!sym(label_col)))
data[[label_col]] <- as.factor(data[[label_col]])

```

Model training.

```{r rf-train-tru-pi}

# Remove columns that are all NA
data <- data[, colSums(is.na(data)) < nrow(data)]

# Make sure all features are numeric except label
num_cols <- sapply(data, is.numeric)
num_cols[label_col] <- FALSE
data <- data[, num_cols | names(data) == label_col]

# Drop columns (genes) with >20% NAs, then impute the rest
na_threshold <- 0.2 * nrow(data)
data_filtered <- data[, colSums(is.na(data)) < na_threshold]

# Impute remaining NAs with median
data_imputed <- data_filtered %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm=TRUE), .)))

# Train Random Forest on all imputed data

# Classes are balanced, 57, 58 samples, no weights needed
rf_model <- ranger(
  x = data_imputed %>% dplyr::select(-all_of(label_col)),
  y = data_imputed[[label_col]],
  num.trees = 500,
  importance = "permutation",
)

# Predict on training data, sanity check
rf_pred_obj <- predict(rf_model, data_imputed %>% dplyr::select(-all_of(label_col)))

# Extract predictions, named list
rf_pred <- rf_pred_obj$predictions

# Make sure predicted and true labels are factors with same levels
rf_pred <- factor(rf_pred, levels = levels(data_imputed[[label_col]]))
true_labels <- factor(data_imputed[[label_col]], levels = levels(data_imputed[[label_col]]))

# Confusion Matrix
confusionMatrix(rf_pred, true_labels)

```

Cross-validation, tune model parameters and save final model.

```{r rf-cv-tru-pi}

cv_folds <- trainControl(method = "cv", number = 5)

# Define mtry values to search over
tune_grid <- expand.grid(
  mtry = seq(30, 70, by = 10),   # centered around 49, sqrt of # features
  splitrule = "gini",
  min.node.size = 5
)

# Cross-validation with grid search using ranger
rf_cv <- train(
  x = data_imputed %>% dplyr::select(-all_of(label_col)),
  y = data_imputed[[label_col]],
  method = "ranger",
  trControl = cv_folds,
  tuneGrid = tune_grid,
  num.trees = 500,
  importance = "permutation", 
)

print(rf_cv)
plot(varImp(rf_cv))

# Retrain model with best mtry found by cv
best_mtry <- rf_cv$bestTune$mtry
best_splitrule <- rf_cv$bestTune$splitrule
best_min_node_size <- rf_cv$bestTune$min.node.size

rf_model_final <- ranger(
  x = data_imputed %>% dplyr::select(-all_of(label_col)),
  y = data_imputed[[label_col]],
  num.trees = 500,
  mtry = best_mtry,
  splitrule = best_splitrule,
  min.node.size = best_min_node_size,
  importance = "permutation",
  classification = TRUE,
)

# Predict on training data, sanity check
rf_pred_obj_final <- predict(rf_model_final, data_imputed %>% dplyr::select(-all_of(label_col)))

# Extract predictions, named list
rf_pred_final <- rf_pred_obj_final$predictions

# Make sure predicted and true labels are factors with same levels
rf_pred_final <- factor(rf_pred_final, levels = levels(data_imputed[[label_col]]))
true_labels <- factor(data_imputed[[label_col]], levels = levels(data_imputed[[label_col]]))

# Confusion Matrix
confusionMatrix(rf_pred_final, true_labels)

saveRDS(rf_model_final, file = "LUAD_Random_Forests/rf_model_final_TRU_vs_proxInflam.rds")
```

#### Feature Importance

Extract most important features for model accuracy, input for annotation step.

```{r rf-tru-pi-feat_imp}

# Extract feature importance from the ranger rf model final
importance_tbl <- rf_model_final$variable.importance

# Convert to data frame, sort descending and scale importance addition
importance_df <- data.frame(
  feature = names(importance_tbl),
  importance = as.numeric(importance_tbl)
) %>%
  arrange(desc(importance)) %>%
  mutate(
    scaled_importance = (importance - min(importance)) / (max(importance) - min(importance))
  )

# Save results
write.csv(importance_df, file = "important_features_TRU_vs_proxInflam.csv", row.names = FALSE)
saveRDS(importance_df, file = "important_features_TRU_vs_proxInflam.rds")

```

```{r rf-tru-pi-feat-imp-plot}

# Feature Importance Plot
library(ggplot2)
top_n <- 20
ggplot(head(importance_df, top_n), aes(x = reorder(feature, importance_df[[1]]), y = importance_df[[1]])) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top Variable Importances", x = "Feature", y = "Importance (Permutation)")
```

### TRU vs proxProlif Binary Classifier Training

This classifier discriminates between subtype classes TRU and proxProlif based on methylation and expression data.

#### Load Feature Matrix and Labels

```{r load-fm-tru-pp}

set.seed(123)

# Set the path to each LUAD training feature matrix
feature_matrix_path <- "Feature Matrices Real Data/feature_matrix_TRU_vs_proxProlif.csv"
label_col <- "subtype"  

data <- read_csv(feature_matrix_path)
str(data)
head(data)

# Remove sample column if present
if ("sample" %in% colnames(data)) {
  data <- data %>% dplyr::select(-sample)
}
# Remove rows where subtype is NA
data <- data %>% filter(!is.na(!!sym(label_col)))
data[[label_col]] <- as.factor(data[[label_col]])

```

Training model.

```{r rf-train-tru-pp}
# Remove columns that are all NA
data <- data[, colSums(is.na(data)) < nrow(data)]

# Make sure all features are numeric except label
num_cols <- sapply(data, is.numeric)
num_cols[label_col] <- FALSE
data <- data[, num_cols | names(data) == label_col]

# Drop columns (genes) with >20% NAs, then impute the rest
na_threshold <- 0.2 * nrow(data)
data_filtered <- data[, colSums(is.na(data)) < na_threshold]

# Impute remaining NAs with median
data_imputed <- data_filtered %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm=TRUE), .)))

# Train Random Forest on all imputed data

# Weighted model to deal with moderate class imbalance
rf_model_w <- ranger(
  x = data_imputed %>% dplyr::select(-all_of(label_col)),
  y = data_imputed[[label_col]],
  num.trees = 500,
  importance = "permutation",
  class.weights = c(TRU = 1, proxProlif = 1.5)
)

rf_pred_obj <- predict(rf_model_w, data_imputed %>% dplyr::select(-all_of(label_col)))

# Extract predictions, named list
rf_pred_w <- rf_pred_obj$predictions

# Make sure predicted and true labels are factors with same levels
rf_pred_w <- factor(rf_pred_w, levels = levels(data_imputed[[label_col]]))
true_labels <- factor(data_imputed[[label_col]], levels = levels(data_imputed[[label_col]]))

# Confusion Matrix
confusionMatrix(rf_pred_w, true_labels)

```

Cross-validation, tune model parameters and save final model.

```{r rf-cv-tru-pp}
cv_folds <- trainControl(method = "cv", number = 5)

# Define mtry values to search over
tune_grid <- expand.grid(
  mtry = c(45, 55, 60, 65, 75),  # centered around sqrt(# features)
  splitrule = "gini",
  min.node.size = 5
)

# Cross-validation with grid search using ranger with class weights
rf_cv <- train(
  x = data_imputed %>% dplyr::select(-all_of(label_col)),
  y = data_imputed[[label_col]],
  method = "ranger",
  trControl = cv_folds,
  tuneGrid = tune_grid,
  num.trees = 500,
  importance = "permutation", 
  weights = ifelse(data_imputed[[label_col]] == "proxProlif", 1.5, 1)
)

print(rf_cv)
plot(varImp(rf_cv))

# Retrain model with best mtry found by cv
best_mtry <- rf_cv$bestTune$mtry
best_splitrule <- rf_cv$bestTune$splitrule
best_min_node_size <- rf_cv$bestTune$min.node.size

rf_model_final <- ranger(
  x = data_imputed %>% dplyr::select(-all_of(label_col)),
  y = data_imputed[[label_col]],
  num.trees = 500,
  mtry = best_mtry,
  splitrule = best_splitrule,
  min.node.size = best_min_node_size,
  importance = "permutation",
  classification = TRUE,
  class.weights = c(TRU = 1, proxProlif = 1.5)
)

# Predict on training data, sanity check
rf_pred_obj_final <- predict(rf_model_final, data_imputed %>% dplyr::select(-all_of(label_col)))

# Extract predictions, named list
rf_pred_final <- rf_pred_obj_final$predictions

# Make sure predicted and true labels are factors with same levels
rf_pred_final <- factor(rf_pred_final, levels = levels(data_imputed[[label_col]]))
true_labels <- factor(data_imputed[[label_col]], levels = levels(data_imputed[[label_col]]))

# Confusion Matrix
confusionMatrix(rf_pred_final, true_labels)

saveRDS(rf_model_final, file = "LUAD_Random_Forests/rf_model_final_TRU_vs_proxProlif.rds")
```

#### Feature Importance

Extract most important features for model accuracy, input for annotation step.

```{r rf-tru-pp-feat_imp}

# Extract feature importance from the ranger rf model final
importance_tbl <- rf_model_final$variable.importance

# Convert to data frame, sort descending and scale importance addition
importance_df <- data.frame(
  feature = names(importance_tbl),
  importance = as.numeric(importance_tbl)
) %>%
  arrange(desc(importance)) %>%
  mutate(
    scaled_importance = (importance - min(importance)) / (max(importance) - min(importance))
  )

# Save results
write.csv(importance_df, file = "important_features_TRU_vs_proxProlif.csv", row.names = FALSE)
saveRDS(importance_df, file = "important_features_TRU_vs_proxProlif.rds")

```

```{r rf-tru-pp-feat-imp-plot}

# Feature Importance Plot
library(ggplot2)
top_n <- 20
ggplot(head(importance_df, top_n), aes(x = reorder(feature, importance_df[[1]]), y = importance_df[[1]])) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top Variable Importances", x = "Feature", y = "Importance (Permutation)")
```

### Random Forest Multi-class Classifier

This classifier was trained using the feature matrix created with the union of the significant genes found in our differential gene expression and methylation analyses.

#### Load Feature Matrix and Labels

```{r load-fm-rf-multiclass}

# Seting a random seed for reproducibility
set.seed(123)

# Create folders for organized output
model_dir <- "RF_results"
if (!dir.exists(model_dir)) dir.create(model_dir)
train_pred_dir <- file.path(model_dir, "Predictions_Train")
if (!dir.exists(train_pred_dir)) dir.create(train_pred_dir)
test_pred_dir <- file.path(model_dir, "Predictions_Test")
if (!dir.exists(test_pred_dir)) dir.create(test_pred_dir)
train_conf_dir <- file.path(model_dir, "ConfusionMatrices_Train")
if (!dir.exists(train_conf_dir)) dir.create(train_conf_dir)
test_conf_dir <- file.path(model_dir, "ConfusionMatrices_Test")
if (!dir.exists(test_conf_dir)) dir.create(test_conf_dir)
train_plots_dir <- file.path(model_dir, "Plots_Train")
if (!dir.exists(train_plots_dir)) dir.create(train_plots_dir)
test_plots_dir <- file.path(model_dir, "Plots_Test")
if (!dir.exists(test_plots_dir)) dir.create(test_plots_dir)
imp_dir <- file.path(model_dir, "Importance")
if (!dir.exists(imp_dir)) dir.create(imp_dir)
bootstrap_dir <- file.path(model_dir, "Bootstrap")
if (!dir.exists(bootstrap_dir)) dir.create(bootstrap_dir)

feature_matrix_path <- "Feature Matrices Real Data/feature_matrix_union_Subtype_Intersected_Genes.csv"
label_col <- "subtype"

data <- read_csv(feature_matrix_path)
if ("sample" %in% colnames(data)) data <- data %>% select(-sample)
data <- data %>% filter(!is.na(!!sym(label_col)))
data[[label_col]] <- as.factor(data[[label_col]])

cat("Class distribution:\n")
print(table(data[[label_col]]))

# Remove all-NA columns
data <- data[, colSums(is.na(data)) < nrow(data)]

# Keep only numeric features + label
num_cols <- sapply(data, is.numeric)
num_cols[label_col] <- FALSE
data <- data[, num_cols | names(data) == label_col]

# Drop columns (genes) with >20% NAs, then impute the rest
na_threshold <- 0.2 * nrow(data)
data_filtered <- data[, colSums(is.na(data)) < na_threshold]

# Impute remaining NAs with median
data_imputed <- data_filtered %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm=TRUE), .)))

```

Multi-class Model Training.

```{r rf-multiclass-train}

# Train multiclass random forest model
rf_model <- randomForest(
  x = data_imputed %>% select(-all_of(label_col)),
  y = data_imputed[[label_col]],
  ntree = 500,
  importance = TRUE
)

# Predict on the training data
rf_pred <- predict(rf_model, data_imputed %>% select(-all_of(label_col)))

# Generate confusion matrix for training data
cm_train <- confusionMatrix(rf_pred, data_imputed[[label_col]])

# Converting confusion matrix to dataframe for saving/plotting
cm_tbl <- as.data.frame(cm_train$table)

# Plotting confusion matrix
p_cm <- ggplot(cm_tbl, aes(Prediction, Reference, fill=Freq)) +
  geom_tile() +
  geom_text(aes(label=Freq), size=6) +
  scale_fill_gradient(low="white", high="steelblue") +
  theme_minimal() +
  labs(title="Confusion Matrix (Train set)", x="Predicted", y="True")

# Save confusion matrix plot
ggsave(file.path(train_plots_dir, "confusion_matrix_train_union_Subtype_Intersected_Genes.png"), plot=p_cm, width=6, height=5)

# Save confusion matrix as CSV and RDS
write.csv(cm_tbl, file.path(train_conf_dir, "confusion_matrix_train_union_Subtype_Intersected_Genes.csv"), row.names=FALSE)
saveRDS(cm_train, file.path(train_conf_dir, "confusion_matrix_train_union_Subtype_Intersected_Genes.rds"))

# Feature Importance Extraction
# Extracting importance scores (MeanDecreaseAccuracy)
importance_tbl <- importance(rf_model, type=1)
importance_df <- as.data.frame(importance_tbl)
importance_df$feature <- rownames(importance_df)

# Sorting by importance descending
importance_df <- importance_df %>% arrange(desc(MeanDecreaseAccuracy))

# Save importance scores as CSV and RDS
write.csv(importance_df, file.path(imp_dir, "important_features_union_Subtype_Intersected_Genes.csv"), row.names=FALSE)
saveRDS(importance_df, file.path(imp_dir, "important_features_union_Subtype_Intersected_Genes.rds"))

# Plotting and save feature importance
png(file.path(train_plots_dir, "varImpPlot_train_union_Subtype_Intersected_Genes.png"), width=800, height=500)
varImpPlot(rf_model, main="Train Feature Importance")
dev.off()
```

Cross-validation, tune multi-class model parameters and save final multi-class model.

```{r rf-multiclass-cv}

# Set up 5-fold cross-validation
cv_folds <- trainControl(method = "cv", number = 5)

# Train Random Forest model with cross-validation to tune 'mtry'
rf_cv <- train(
  x = data_imputed %>% select(-all_of(label_col)),
  y = data_imputed[[label_col]],
  method = "rf",
  trControl = cv_folds,
  ntree = 500,
  importance = TRUE
)

# Print cross-validation summary (accuracy per mtry value)
print(rf_cv)

# Extract the best 'mtry' value (number of features considered at each split)
best_mtry <- rf_cv$bestTune$mtry

# Retrain the final model using the best 'mtry'
rf_model_final <- randomForest(
  x = data_imputed %>% select(-all_of(label_col)),  
  y = data_imputed[[label_col]],                   
  ntree = 500,
  mtry = best_mtry, # Use best mtry found by CV
  importance = TRUE
)

# Save the final trained model
saveRDS(rf_model_final, file.path(model_dir, "rf_model_final_union_Subtype_Intersected_Genes.rds"))
```

## 2.6.2 Support Vector Machines - Training, Tuning, Evaluation

### Binary SVM Classifiers - Train & Test Data 

Load feature matrices for both test and training sets. \
\
**NOTE:** For practicality, we have automated the generation of SVM results, including testing the models on the test set.

```{r svm-fm-paths}

train_dir <- "Feature Matrices Real Data"
test_dir <- "Test Data/LUAD_feature_matrices"

model_dir <- "SVM_results"
dir.create(model_dir, showWarnings=FALSE)
pred_dir <- file.path(model_dir, "Predictions"); dir.create(pred_dir, showWarnings=FALSE)
conf_dir <- file.path(model_dir, "ConfusionMatrices"); dir.create(conf_dir, showWarnings=FALSE)
imp_dir <- file.path(model_dir, "Importance"); dir.create(imp_dir, showWarnings=FALSE)
plots_dir <- file.path(model_dir, "Plots"); dir.create(plots_dir, showWarnings=FALSE)

train_files <- list.files(train_dir, pattern = "\\.csv$", full.names = TRUE)
results_list <- list()
```

SVM models training, model tuning, saving final mode, final feature importance extraction and bootstraping.

```{r svm-train}
for (train_path in train_files) {
  comp <- tools::file_path_sans_ext(basename(train_path)) %>% gsub("^feature_matrix_", "", .)
  test_path <- file.path(test_dir, paste0("test_input_matrix_", comp, ".csv"))

  # Load data
  train <- read_csv(train_path)
  test  <- read_csv(test_path)

  if ("sample" %in% colnames(train)) train <- train %>% select(-sample)
  train <- train %>% filter(!is.na(subtype))
  train$subtype <- as.factor(train$subtype)
  if ("sample" %in% colnames(test)) test <- test %>% select(-sample)
  test <- test %>% filter(!is.na(subtype))
  test$subtype <- as.factor(test$subtype)

  X <- train %>% select(-subtype)
  y <- train$subtype
  Xtest <- test %>% select(-subtype)
  ytest <- test$subtype

  for (col in colnames(X)) X[[col]][is.na(X[[col]])] <- median(X[[col]], na.rm=TRUE)
  for (col in colnames(Xtest)) if (col %in% colnames(X)) Xtest[[col]][is.na(Xtest[[col]])] <- median(X[[col]], na.rm=TRUE)

  keep_cols <- sapply(X, function(col) !(all(is.na(col)) | sd(col, na.rm=TRUE) == 0))
  X <- X[, keep_cols, drop=FALSE]
  Xtest <- Xtest[, colnames(X), drop=FALSE]

  # Feature selection: top 200 features by variance (if many)
  if (ncol(X) > 200) {
    vars <- apply(X, 2, var)
    top_vars <- order(vars, decreasing = TRUE)[1:200]
    X <- X[, top_vars, drop=FALSE]
    Xtest <- Xtest[, colnames(X), drop=FALSE]
  }

  # Class weights for imbalance
  wtable <- table(y)
  class_weights <- as.numeric(max(wtable) / wtable)
  names(class_weights) <- names(wtable)

  cat("Class distribution:", paste(names(wtable), wtable, collapse = " "), "\n")

  # Grid search for C
  c_grid <- c(0.01, 0.1, 1, 10, 100)
  tune_result <- tryCatch({
    tune(
      svm,
      train.x = X,
      train.y = y,
      kernel = "linear",
      ranges = list(cost = c_grid),
      class.weights = class_weights,
      scale = FALSE,
      probability = TRUE,
      cross = 3
    )
  }, error = function(e) {
    cat("Grid search failed:\n"); print(e)
    return(NULL)
  })

  if (is.null(tune_result) || is.null(tune_result$best.parameters$cost)) {
    cat("Grid search failed or cost not found, using cost=1 by default.\n")
    best_C <- 1
  } else {
    best_C <- tune_result$best.parameters$cost
    cat("Best cost found:", best_C, "\n")
  }

  # Fit final model
  svm_final <- svm(
    x = X,
    y = y,
    kernel = "linear",
    cost = best_C,
    class.weights = class_weights,
    scale = FALSE,
    probability = TRUE
  )
  saveRDS(svm_final, file=file.path(model_dir, paste0("svm_model_", comp, ".rds")))

  # Predict and evaluate
  pred <- predict(svm_final, Xtest)
  conf <- caret::confusionMatrix(pred, ytest)
  write_csv(tibble(True=ytest, Pred=pred), file=file.path(pred_dir, paste0("predictions_", comp, ".csv")))
  saveRDS(conf, file=file.path(conf_dir, paste0("confMat_", comp, ".rds")))

  # Confusion matrix plot
  cm_table <- as.data.frame(conf$table)
  p_cm <- ggplot(cm_table, aes(x=Prediction, y=Reference, fill=Freq)) +
    geom_tile() + geom_text(aes(label=Freq), size=6) +
    scale_fill_gradient(low="white", high="steelblue") + theme_minimal() +
    labs(title=paste("Confusion Matrix -", comp), x="Predicted", y="True")
  ggsave(file.path(plots_dir, paste0("confusion_matrix_", comp, ".png")), plot=p_cm, width=5, height=4)

  # Feature importance (t-test/ANOVA)
  if (nlevels(y) == 2) {
    imp_vec <- sapply(X, function(x) abs(t.test(x ~ y)$statistic))
  } else {
    imp_vec <- sapply(X, function(x) summary(aov(x ~ y))[[1]][["F value"]][1])
  }
  imp_vec <- sort(imp_vec, decreasing=TRUE)
  imp_df <- data.frame(Feature = names(imp_vec), Importance = as.numeric(imp_vec))
  write_csv(imp_df, file.path(imp_dir, paste0("feature_importance_", comp, ".csv")))

  # Bootstrapping
  n_boot <- 1000
  accs <- numeric(n_boot)
  sens <- numeric(n_boot)
  spec <- numeric(n_boot)
  kappas <- numeric(n_boot)
  for (i in 1:n_boot) {
    idx <- sample(seq_along(ytest), replace=TRUE)
    yb <- ytest[idx]
    xb <- Xtest[idx, , drop=FALSE]
    pb <- predict(svm_final, xb)
    cm <- caret::confusionMatrix(pb, yb)
    accs[i] <- cm$overall["Accuracy"]
    if (is.matrix(cm$byClass)) {
      sens[i] <- mean(cm$byClass[,"Sensitivity"], na.rm=TRUE)
      spec[i] <- mean(cm$byClass[,"Specificity"], na.rm=TRUE)
    } else {
      sens[i] <- cm$byClass["Sensitivity"]
      spec[i] <- cm$byClass["Specificity"]
    }
    kappas[i] <- cm$overall["Kappa"]
  }
  # Save bootstrapping results
  boot_df <- tibble(
    accuracy = accs, sensitivity = sens, specificity = spec, kappa = kappas
  )
  write_csv(boot_df, file=file.path(model_dir, paste0("bootstrap_results_", comp, ".csv")))
  saveRDS(boot_df, file=file.path(model_dir, paste0("bootstrap_results_", comp, ".rds")))

  # Bootstrapping plots
  p_hist <- ggplot(boot_df, aes(x=accuracy)) +
    geom_histogram(bins=30, fill="#1f77b4", color="white") +
    theme_minimal() +
    labs(title=paste("Bootstrap Accuracy -", comp), x="Accuracy", y="Count")
  ggsave(file.path(plots_dir, paste0("bootstrap_hist_accuracy_", comp, ".png")), plot=p_hist, width=6, height=4)

  boot_long <- boot_df %>% pivot_longer(everything(), names_to="Metric", values_to="Value")
  p_box <- ggplot(boot_long, aes(x=Metric, y=Value, fill=Metric)) +
    geom_boxplot(alpha=0.7) + theme_minimal() +
    scale_fill_brewer(palette="Set2") +
    labs(title=paste("Bootstrap Metrics -", comp))
  ggsave(file.path(plots_dir, paste0("bootstrap_boxplot_metrics_", comp, ".png")), plot=p_box, width=6, height=4)

  results_list[[comp]] <- list(
    confusion = conf, accs = accs, sens = sens, spec = spec,
    kappas = kappas, best_C = best_C, class_weights = class_weights
  )
}
```

### Multi-class SVM Classifier

This model will be trained using the union of the gene intersections of our significant differentially expressed and methylated genes.\
\
**NOTE:** For practicality, we have automated the generation of SVM results, including testing the models on the test set.

```{r svm-multiclass-paths}

model_dir <- "SVM_results_multiclass"
dir.create(model_dir, showWarnings=FALSE)
pred_dir <- file.path(model_dir, "Predictions"); dir.create(pred_dir, showWarnings=FALSE)
conf_dir <- file.path(model_dir, "ConfusionMatrices"); dir.create(conf_dir, showWarnings=FALSE)
imp_dir <- file.path(model_dir, "Importance"); dir.create(imp_dir, showWarnings=FALSE)
plots_dir <- file.path(model_dir, "Plots"); dir.create(plots_dir, showWarnings=FALSE)
bootstrap_dir <- file.path(model_dir, "Bootstrap"); dir.create(bootstrap_dir, showWarnings=FALSE)

train_path <- "Feature Matrices Real Data/feature_matrix_union_Subtype_Intersected_Genes.csv"
test_path <- "Test Data/LUAD_feature_matrices/test_input_matrix_union_Subtype_Intersected_Genes.csv"
comp <- "union_Subtype_Intersected_Genes"

# Load data
train <- read_csv(train_path)
test  <- read_csv(test_path)

if ("sample" %in% colnames(train)) train <- train %>% select(-sample)
train <- train %>% filter(!is.na(subtype))
train$subtype <- as.factor(train$subtype)
if ("sample" %in% colnames(test)) test <- test %>% select(-sample)
test <- test %>% filter(!is.na(subtype))
test$subtype <- as.factor(test$subtype)

X <- train %>% select(-subtype)
y <- train$subtype
Xtest <- test %>% select(-subtype)
ytest <- test$subtype

# Impute NAs (median for numeric columns)
for (col in colnames(X)) X[[col]][is.na(X[[col]])] <- median(X[[col]], na.rm=TRUE)
for (col in colnames(Xtest)) if (col %in% colnames(X)) Xtest[[col]][is.na(Xtest[[col]])] <- median(X[[col]], na.rm=TRUE)

# Remove all-NA and zero-variance cols
keep_cols <- sapply(X, function(col) !(all(is.na(col)) | sd(col, na.rm=TRUE) == 0))
X <- X[, keep_cols, drop=FALSE]
Xtest <- Xtest[, colnames(X), drop=FALSE]
```

Multi-class SVM model training, parameter tuning, bootstrapping, saving final model, and feature importance extraction for annotation.

```{r svm-multiclass-train}

# Feature selection: top 200 features by variance (if many)
if (ncol(X) > 200) {
  vars <- apply(X, 2, var)
  top_vars <- order(vars, decreasing = TRUE)[1:200]
  X <- X[, top_vars, drop=FALSE]
  Xtest <- Xtest[, colnames(X), drop=FALSE]
}

# Class weights for imbalance (works for multiclass)
wtable <- table(y)
class_weights <- as.numeric(max(wtable) / wtable)
names(class_weights) <- names(wtable)
cat("Class distribution:", paste(names(wtable), wtable, collapse = " "), "\n")

# Grid search for C 
c_grid <- c(0.01, 0.1, 1, 10, 100)
tune_result <- tryCatch({
  tune(
    svm,
    train.x = X,
    train.y = y,
    kernel = "linear",
    ranges = list(cost = c_grid),
    class.weights = class_weights,
    scale = FALSE,
    probability = TRUE,
    cross = 3
  )
}, error = function(e) {
  cat("Grid search failed:\n"); print(e)
  return(NULL)
})

if (is.null(tune_result) || is.null(tune_result$best.parameters$cost)) {
  cat("Grid search failed or cost not found, using cost=1 by default.\n")
  best_C <- 1
} else {
  best_C <- tune_result$best.parameters$cost
  cat("Best cost found:", best_C, "\n")
}

# Fit final model 
svm_final <- svm(
  x = X,
  y = y,
  kernel = "linear",
  cost = best_C,
  class.weights = class_weights,
  scale = FALSE,
  probability = TRUE
)
saveRDS(svm_final, file=file.path(model_dir, paste0("svm_model_", comp, ".rds")))

# Predict and evaluate 
pred <- predict(svm_final, Xtest)
conf <- caret::confusionMatrix(pred, ytest)
write_csv(tibble(True=ytest, Pred=pred), file=file.path(pred_dir, paste0("predictions_", comp, ".csv")))
saveRDS(conf, file=file.path(conf_dir, paste0("confMat_", comp, ".rds")))

# Confusion matrix plot
cm_table <- as.data.frame(conf$table)
p_cm <- ggplot(cm_table, aes(x=Prediction, y=Reference, fill=Freq)) +
  geom_tile() + geom_text(aes(label=Freq), size=6) +
  scale_fill_gradient(low="white", high="steelblue") + theme_minimal() +
  labs(title=paste("Confusion Matrix -", comp), x="Predicted", y="True")
ggsave(file.path(plots_dir, paste0("confusion_matrix_", comp, ".png")), plot=p_cm, width=5, height=4)

# Feature importance: multiclass via ANOVA F-value 
imp_vec <- sapply(X, function(x) summary(aov(x ~ y))[[1]][["F value"]][1])
imp_vec <- sort(imp_vec, decreasing=TRUE)
imp_df <- data.frame(Feature = names(imp_vec), Importance = as.numeric(imp_vec))
write_csv(imp_df, file.path(imp_dir, paste0("feature_importance_", comp, ".csv")))

# Bootstrapping (test set metrics) 
n_boot <- 1000
accs <- numeric(n_boot)
sens <- numeric(n_boot)
spec <- numeric(n_boot)
kappas <- numeric(n_boot)
for (i in 1:n_boot) {
  idx <- sample(seq_along(ytest), replace=TRUE)
  yb <- ytest[idx]
  xb <- Xtest[idx, , drop=FALSE]
  pb <- predict(svm_final, xb)
  cm <- caret::confusionMatrix(pb, yb)
  accs[i] <- cm$overall["Accuracy"]
  # Sens/spec: average all classes (multiclass)
  sens[i] <- mean(cm$byClass[,"Sensitivity"], na.rm=TRUE)
  spec[i] <- mean(cm$byClass[,"Specificity"], na.rm=TRUE)
  kappas[i] <- cm$overall["Kappa"]
}
# Save bootstrapping results
boot_df <- tibble(
  accuracy = accs, sensitivity = sens, specificity = spec, kappa = kappas
)
write_csv(boot_df, file=file.path(bootstrap_dir, paste0("bootstrap_results_", comp, ".csv")))
saveRDS(boot_df, file=file.path(bootstrap_dir, paste0("bootstrap_results_", comp, ".rds")))

# Bootstrapping plots 
p_hist <- ggplot(boot_df, aes(x=accuracy)) +
  geom_histogram(bins=30, fill="#1f77b4", color="white") +
  theme_minimal() +
  labs(title=paste("Bootstrap Accuracy -", comp), x="Accuracy", y="Count")
ggsave(file.path(plots_dir, paste0("bootstrap_hist_accuracy_", comp, ".png")), plot=p_hist, width=6, height=4)

boot_long <- boot_df %>% pivot_longer(everything(), names_to="Metric", values_to="Value")
p_box <- ggplot(boot_long, aes(x=Metric, y=Value, fill=Metric)) +
  geom_boxplot(alpha=0.7) + theme_minimal() +
  scale_fill_brewer(palette="Set2") +
  labs(title=paste("Bootstrap Metrics -", comp))
ggsave(file.path(plots_dir, paste0("bootstrap_boxplot_metrics_", comp, ".png")), plot=p_box, width=6, height=4)

# Save summary in results_list
results_list <- list()
results_list[[comp]] <- list(
  confusion = conf, accs = accs, sens = sens, spec = spec,
  kappas = kappas, best_C = best_C, class_weights = class_weights
)
```

After Data Modeling, we have produced, trained and tuned binary Random Forest and SVM Classifiers, as well as Multi-class Random Forest and SVM classifiers.
