---

title: "LUAD Multi-Omics Classification: Model Evaluation"
output:
html_document:
df_print: paged
---------------

# Model Evaluation

## Project Overview

This notebook evaluates the performance of machine learning models trained to classify LUAD molecular subtypes using integrated transcriptomic and methylation features.

Performance is assessed on an independent test dataset using multiple evaluation metrics.

## Pipeline Position

Step 3 of 4

Data Acquisition → Data Integration → Feature Selection & Modeling → Evaluation → Biological Interpretation

## Evaluation Metrics

* Accuracy
* Sensitivity
* Specificity
* Cohen's Kappa
* Bootstrap validation

## Outputs

* Confusion matrices
* Classification metrics
* Bootstrap performance estimates
* Comparative model summaries

## Authors

Developed as part of a collaborative academic project at Freie Universität Berlin.


# Test Data

Our final project focused on working with the TCGA LUAD (Lung Adenocarcinoma) RNA-Seq, Methylation and Clinical data. This R Notebook contains the code corresponding to the Test Data processing and model evaluation by predicting labels on unseen data using the models trained in the data modeling step.

The test data processing includes the following sections:

a.  RNA-Seq & Methylation Test Data Sample Matching
b.  Test Data Normalization
c.  Test Data Input Matrix Preparation
d.  Model Prediction on Test Data

## Setup

Load required libraries.

```{r setup, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(recount3)
library(SummarizedExperiment)
library(tidyverse)
library(DESeq2)
library(org.Hs.eg.db)
library(caret)

```

## RNA-Seq & Methylation Test Data Sample Matching

The test data will undergo the same preprocessing steps as the training data, to ensure we provide the correct format to our trained models.

```{r test-data-sample-match}

# Load RSE object with split labels

# This dataset was pre-processed by filtering:
# Sample: Only primary and normal kept
# Gene: Only protein coding kept
# Counts: Only genes with 10 counts on at least 10 samples kept
# And then we carried out a stratified split into 80% train, 20% test
load("rse_objects/rse_luad_sgc_filtered.RData")

# Load methylation data
load("met_data_files/LUAD_methylation_SE.rda")

# Filter out MASK_general == TRUE probes, as these are bad quality
luad_met_filtered <- luad_met[!rowData(luad_met)$MASK_general %in% TRUE, ]

# Extract β-matrix and metadata
beta_mat  <- assay(luad_met_filtered)
meta_meth <- as.data.frame(colData(luad_met_filtered))

# Subset to RNA‐seq samples only
rse_test <- rse_luad_sgc_filtered[, colData(rse_luad_sgc_filtered)$split == "test"]
rna_ids   <- substr(colData(rse_test)$tcga.cgc_sample_id, 1, 16)
meth_ids  <- substr(colnames(beta_mat), 1, 16)
common <- intersect(rna_ids, meth_ids) 

# Select only common samples of type primary tumor
primary_tumor_met_samples <- meta_meth$sample[meta_meth$sample_type == "Primary Tumor"]

# Intersect RNA-matched methylation samples with primary tumors only
common_primary <- intersect(common, primary_tumor_met_samples) 

# Align and subset the β‐matrix with only common primary tumor samples
idx_beta <- match(common_primary, meth_ids)
beta_sub <- beta_mat[, idx_beta]

# Filter RNA-seq rse_test to keep only matching primary tumor samples
rna_ids_primary <- match(common_primary, rse_test$tcga.cgc_sample_id)
rse_test_primary <- rse_test[, rna_ids_primary]

# Create gene annotation vector using luad_met metadata
gene_annot_vector <- rowData(luad_met_filtered)$gene_HGNC #421433, corresponds to genes mapped to probes
names(gene_annot_vector) <- rownames(luad_met_filtered)

```

## Test Data Normalisation

Our trained models expect normalised data, so the test data will be normalised in the same way as the training data: the expression values will be normalized using the *DESeq2::vst* function, and the methylation values will be transformed to M values.

```{r test-data-norm, message=FALSE}

# Models require normalized expression counts and methylation M values.

# Methylation data 
# Convert beta values to M-values
percentM <- beta_sub * 100
percentM <- pmin(pmax(percentM, 0.01), 99.99)  # clamp to avoid ±Inf
Mval <- log2(percentM / (100 - percentM))

# RNA-seq data
# Build factor in same order as rse_test_primary
# Subtype
subtypes_test <- factor(
  rse_test_primary$expression_subtype[match(common_primary, rse_test_primary$tcga.cgc_sample_id)],
  levels = c("prox.-inflam", "prox.-prolif.", "TRU"),
  labels = c("proxInflam", "proxProlif", "TRU") # for readability
)

# Subtype NA filtering
keep_subtype <- !is.na(subtypes_test)
Mval_subtype <- Mval[, keep_subtype] 
subtypes_test_filtered <- subtypes_test[keep_subtype]

# Subset rse_train_primary to only contain samples with subtype
short_ids <- stringr::str_sub(rse_test_primary$tcga.cgc_sample_id, 1, 12)
rse_test_primary$short_id <- short_ids
keep_samples <- short_ids %in% names(subtypes_test_filtered)
rse_test_subtyped <- rse_test_primary[, keep_samples]

# RNA-seq data
raw_counts <- assays(rse_test_subtyped)$raw_counts
vst_obj <- vst(DESeqDataSetFromMatrix(
  countData = raw_counts,
  colData   = colData(rse_test_subtyped),
  design    = ~ 1
))

expr_mat <- assay(vst_obj)  # vst-normalized expression

# Assign TCGA sample ids as colnames for expr_mat
colnames(expr_mat) <- rse_test_subtyped$tcga.cgc_sample_id

```

## Test Data Input Matrix Preparation

The test data will be formatted as the train data for label prediction.

```{r rna-data-processing, message=FALSE}

# RNA-seq data 

# Ensemble to HGNC gene ID mapping

# ensembl_to_symbol function: convert RNA-seq gene Ensemble ids to HGNC, returns a named vector

ensembl_to_symbol <- function(ensembl_ids) {
  ids_noversion <- sub("\\..*$", "", ensembl_ids)
  
  map <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys    = unique(ids_noversion),
    keytype = "ENSEMBL",
    columns = "SYMBOL"
  )
  
  map <- map[!is.na(map$SYMBOL) & map$SYMBOL != "", ]
  map <- map[!duplicated(map$ENSEMBL), ]

  # Named vector: names = Ensembl, values = SYMBOL
  hgnc_vec <- setNames(map$SYMBOL, map$ENSEMBL)

  # Return in input order
  return(hgnc_vec[ids_noversion])
}

# Map ENSEMBLE gene IDs to HGNC and average duplicates
# Multiple ensemble IDs map to HGNC symbols, so two or more rows can have the same row name (HGNC gene) with different expression values which will be averaged to keep only one row. 
ensembl_ids <- gsub("\\..*", "", rownames(expr_mat))  # Remove version
hgnc_names <- ensembl_to_symbol(ensembl_ids)
gene_symbols <- hgnc_names[ensembl_ids]

# Add gene symbols as a column to a df before assigning as rownames
expr_df <- as.data.frame(expr_mat)
expr_df$gene_symbol <- gene_symbols

# Move gene_symbol to the first column
expr_df <- expr_df[, c(ncol(expr_df), 1:(ncol(expr_df) - 1))]

#Convert to tibble
expr_df <- as_tibble(expr_df)

# Filter NA gene symbols and aggregate duplicated gene expression values
expr_df <- expr_df %>%
  filter(!is.na(gene_symbol) & gene_symbol != "")

# Average duplicates
expr_avg <- expr_df %>%
  group_by(gene_symbol) %>%
  summarise(across(everything(), mean), .groups = "drop")

# Replace expr_mat with expr_avg to use downstream as matrix again
expr_mat <- expr_avg %>%
  column_to_rownames("gene_symbol") %>%
  as.matrix()

```

```{r meth-data-processing}

# Methylation data

# Mapping CpG probes to gene symbols
# Collapsing multiple CpG probes that map to one gene. 

# annot_tbl, one probe per row mapped to corresponding gene
# probes mapping to more than one gene get their own row
annot_tbl <- tibble(
  probe = names(gene_annot_vector),
  gene  = gene_annot_vector
) %>%
  separate_rows(gene, sep = ";") %>%
  filter(!is.na(gene) & gene != "") %>%
  filter(probe %in% rownames(Mval_subtype))

# mval_long, get gene level M values
mval_long <- Mval_subtype[annot_tbl$probe, , drop = FALSE] %>%
  as.data.frame() %>%
  rownames_to_column("probe") %>%
  left_join(annot_tbl, by = "probe") %>%
  filter(!is.na(gene) & gene != "") %>%
  pivot_longer(-c(probe, gene), names_to = "sample", values_to = "mval") %>%
  group_by(gene, sample) %>%
  summarise(meth = mean(mval, na.rm = TRUE), .groups = "drop")

#methylation values matrix, rows = genes, columns = samples
meth_mat <- mval_long %>%
  pivot_wider(names_from = sample, values_from = meth) %>%
  column_to_rownames("gene") %>%
  as.matrix()
#33123 genes, 40 samples with M values

# Shortening sample names to match expression matrix
colnames(meth_mat) <- substr(colnames(meth_mat), 1, 16)

```

```{r combined-matrix-processing}

# Intersect genes present in both omics, to create a joint matrix
common_genes <- intersect(rownames(expr_mat), rownames(meth_mat)) #17670 common genes
expr_mat_sub <- expr_mat[common_genes, ]
meth_mat_sub <- meth_mat[common_genes, ]

expr_mat <- expr_mat_sub
meth_mat <- meth_mat_sub

# Matrix transposing, scaling works when features are cols
expr_mat <- t(expr_mat)
meth_mat <- t(meth_mat)

# Keep omics type
colnames(expr_mat) <- paste0("expr_", colnames(expr_mat))
colnames(meth_mat) <- paste0("meth_", colnames(meth_mat))

# Combine both matrix and standardize omics values (scale)
combined_mat <- cbind(expr_mat, meth_mat)

```

The combined matrix containing expresion and methylation values will be scaled using the same mean and standard deviation as for the training data to avoid bias.

```{r test-data-scaling}

# Load means and sds used for scaling train data
train_means <- readRDS("feature_matrices/train_scaling_means.rds")
train_sds <- readRDS("feature_matrices/train_scaling_sds.rds")

# Scale matrix using training means and sds
scaled_combined_test_mat <- sweep(combined_mat, 2, train_means, FUN = "-")
scaled_combined_test_mat <- sweep(scaled_combined_test_mat, 2, train_sds, FUN = "/")

```

```{r save-scaled-test-matrix}

# Create "feature_matrices" folder if it does not exist
test_data_dir <- "test_data"
if (!dir.exists(test_data_dir)) dir.create(test_data_dir)

# Save test scaled combined matrix in test_data dir
saveRDS(scaled_combined_test_mat, 
        file = file.path(test_data_dir, "LUAD_test_scaled_expr_meth_mat.rds"))

```

The test input matrices for prediction with our trained models will be created from the combined scaled test matrix in the same way as for the training data.

```{r test-input-matrices}

# We have 3 LUAD binary classification models, 1 per subtype pairwise comparison
# Each pairwise comparison has an intersection of DEG & DMPs significant genes
# Build input test data matrices for each model using the intersection genes and also the union for multi-class classifier

inter_dir <- "intersections"
out_dir <- "feature_matrices"
log_file <- "feature_matrices/test_input_matrix_log.txt"
dir.create(out_dir, showWarnings = FALSE)

# Open log connection
log_con <- file(log_file, open = "wt")
sink(log_con, type = "output")

# List gene intersection and union files
inter_files <- list.files(inter_dir, full.names = TRUE)

for (f in inter_files) {
  comp <- tools::file_path_sans_ext(basename(f)) %>%
    gsub("^Intersection_", "", .) %>%
    gsub("^Union_", "union_", .)
  
  cat("\n>>> Processing:", comp, "\n")

  genes <- read_csv(f, show_col_types = FALSE)$Gene %>% toupper()
  
  # Match columns in combined matrix (columns look like expr_X or meth_X)
  expr_cols <- paste0("expr_", genes)
  meth_cols <- paste0("meth_", genes)
  gene_keep <- intersect(colnames(scaled_combined_test_mat), c(expr_cols, meth_cols))
  
  cat(" - Genes in input file:", length(genes), "\n")
  cat(" - Matched columns in combined matrix:", length(gene_keep), "\n")
  
  if (length(gene_keep) < 5) {
    cat(" !! Too few matched columns for", comp, "\n")
    next
  }
  
  # Subset feature matrix by genes
  sub_df <- scaled_combined_test_mat[, gene_keep, drop = FALSE]
  sub_df <- as.data.frame(sub_df)
  
  # Match sample short IDs to subtype mapping
  sub_df$sample <- rownames(sub_df)
  sub_df$short_id <- substr(sub_df$sample, 1, 12)
  sub_df$subtype <- subtypes_test_filtered[sub_df$short_id]
  
  # Filter to keep only labeled samples
  sub_df <- sub_df[!is.na(sub_df$subtype), ]
  
  # For intersections: filter to the two subtypes for pairwise comparison
  if (grepl("^Intersection_", basename(f))) {
    subtype_pair <- unlist(strsplit(comp, "_vs_"))
    sub_df <- sub_df[sub_df$subtype %in% subtype_pair, ]

    if (length(unique(sub_df$subtype)) != 2) {
      cat(" !! Warning: expected 2 subtypes, found:", paste(unique(sub_df$subtype), collapse = ", "), "\n")
    }
  }
  
  # Clean up: drop short_id, reorder
  sub_df$short_id <- NULL
  sub_df <- sub_df[, c(setdiff(colnames(sub_df), c("sample", "subtype")), "sample", "subtype")]

  cat(" - Samples:", nrow(sub_df), "\n")
  cat(" - Features (expr + meth):", length(gene_keep), "\n")
  cat(" - Total columns (incl. sample + label):", ncol(sub_df), "\n")

  # Save final matrix
  write_csv(sub_df, file.path(out_dir, paste0("test_input_matrix_", comp, ".csv")))
  saveRDS(sub_df, file.path(out_dir, paste0("test_input_matrix_", comp, ".rds")))
  cat(" - Saved to:", file.path(out_dir, paste0("test_input_matrix_", comp, ".csv")), "\n")
}

# Close log
sink()
close(log_con)

```

## Model Prediction on Test Data

Our trained RF models will be loaded to be applied for label prediction of our test data, and the results evaluated. We carried out predictions for the binary classifiers and for multi-class classifiers. \
\
**NOTE:** SVM binary and multi-class predictions and results can be found in the 2_LUAD_DataModeling.Rmd notebook, section 2.6.2. As explained in that section, and after our experience with the random forest model training and predictions, we decided to optimise and automate the SVM training, tuning and prediction for practicality.

### Random Forest Binary Classifiers

#### proxProlif vs proxInflam Subtype Prediction

```{r model-pred-pp-pi}

# Model Prediction on Test data

# proxProlif vs proxInflam binary classification
# Load input data

# Set the path to proxProlif vs proxInflam test data
test_matrix_path <- "LUAD_Test_Data/test_input_matrix_proxProlif_vs_proxInflam.csv"
# True label col, for model evaluation on unseen data
label_col <- "subtype"

data <- read_csv(test_matrix_path)
str(data)
head(data)

# Remove sample column if present
if ("sample" %in% colnames(data)) {
  data <- data %>% dplyr::select(-sample)
}

# Remove rows where subtype is NA
data <- data %>% filter(!is.na(!!sym(label_col)))
data[[label_col]] <- as.factor(data[[label_col]])

# Remove columns that are all NA
data <- data[, colSums(is.na(data)) < nrow(data)]

# Make sure all features are numeric except label
num_cols <- sapply(data, is.numeric)
num_cols[label_col] <- FALSE
data <- data[, num_cols | names(data) == label_col]

# Drop columns (genes) with >20% NAs, then impute the rest
na_threshold <- 0.2 * nrow(data)
data_filtered <- data[, colSums(is.na(data)) < na_threshold]

# Impute remaining NAs with median
data_imputed <- data_filtered %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm=TRUE), .)))

# Load model
rf_model_pp_pi <- readRDS("LUAD_Random_Forests/rf_model_final_proxProlif_vs_proxInflam.rds")

# Predict labels
rf_pred_pp_pi_obj <- predict(rf_model_pp_pi, data_imputed %>% dplyr::select(-all_of(label_col)))

# Extract predictions, named list
rf_pred_pp_pi <- rf_pred_pp_pi_obj$predictions

# Make sure predicted and true labels are factors with same levels
rf_pred_pp_pi <- factor(rf_pred_pp_pi, levels = levels(data_imputed[[label_col]]))
true_labels <- factor(data_imputed[[label_col]], levels = levels(data_imputed[[label_col]]))

# Model evaluation

# Confusion Matrix using true labels
confMat_pp_pi <- confusionMatrix(rf_pred_pp_pi, data_imputed[[label_col]])
# Accuracy: 91.6%, pretty good! 
# Saving confusion matrix to revisit if needed, metrics are here
saveRDS(confMat_pp_pi, file = "confMat_pp_pi.rds")


# Bootstrapping, check reliability of our results
# Useful for our small test sample size of 24

# Separating input features and labels from test data
X_test <- data_imputed %>% dplyr::select(-all_of(label_col))
y_test <- data_imputed[[label_col]]

# Specify number of bootstrap samples
n_boot <- 1000

# Store metrics per boostrap in a vector
accuracies <- numeric(n_boot)
sensitivities <- numeric(n_boot)
specificities <- numeric(n_boot)
kappas <- numeric(n_boot)

# Assign model here to variable name used by for
rf_model <- rf_model_pp_pi

# Seed for reproducibility
set.seed(145)  

for (i in 1:n_boot) {
  # Sample indices with replacement
  sample_idx <- sample(seq_len(nrow(X_test)), replace = TRUE)
  
  # Create bootstrap sample
  X_boot <- X_test[sample_idx, ]
  y_boot <- y_test[sample_idx]
  
  # Prediction
  pred_boot_obj <- predict(rf_model, X_boot)
  pred_boot <- factor(pred_boot_obj$predictions, levels = levels(y_boot))  # force factor with correct levels
  
  # Evaluation
  cm <- confusionMatrix(pred_boot, y_boot)
  accuracies[i] <- cm$overall["Accuracy"]
  sensitivities[i] <- cm$byClass["Sensitivity"]
  specificities[i] <- cm$byClass["Specificity"]
  kappas[i] <- cm$overall["Kappa"]
}

# Summary of metrics
summary(accuracies)
summary(sensitivities)
summary(specificities)
summary(kappas)

# Overall boostrap results from summary observations for proxProlif vs proxInflam RF
# High accuracy: ~91%
# High sensitivity for proxInflam (TP for proxInflam) ~93%
# Decent specificity (TN, pretty good at detecting proxProlif) ~89%
# Kappa, agreement beyond chance is robust ~81%

# Metric df to save
bootstrap_results_df_pp_pi <- data.frame(
  accuracy = accuracies,
  sensitivity = sensitivities,
  specificity = specificities,
  kappa = kappas
)

# Save metrics df
write.csv(bootstrap_results_df_pp_pi, "bootstrap_results_proxProlif_vs_proxInflam.csv", row.names = FALSE)
saveRDS(bootstrap_results_df_pp_pi, file = "bootstrap_results_proxProlif_vs_proxInflam.rds")

```

#### **TRU vs proxInflam Subtype Prediction**

```{r model-pred-tru-pi}

# TRU vs proxInflam binary classification
# Load input data

# Set the path to TRU vs proxInflam test data
test_matrix_path <- "LUAD_Test_Data/test_input_matrix_TRU_vs_proxInflam.csv"
# True label col, for model evaluation on unseen data
label_col <- "subtype"

data <- read_csv(test_matrix_path)
str(data)
head(data)

# Remove sample column if present
if ("sample" %in% colnames(data)) {
  data <- data %>% dplyr::select(-sample)
}

# Remove rows where subtype is NA
data <- data %>% filter(!is.na(!!sym(label_col)))
data[[label_col]] <- as.factor(data[[label_col]])

# Remove columns that are all NA
data <- data[, colSums(is.na(data)) < nrow(data)]

# Make sure all features are numeric except label
num_cols <- sapply(data, is.numeric)
num_cols[label_col] <- FALSE
data <- data[, num_cols | names(data) == label_col]

# Drop columns (genes) with >20% NAs, then impute the rest
na_threshold <- 0.2 * nrow(data)
data_filtered <- data[, colSums(is.na(data)) < na_threshold]

# Impute remaining NAs with median
data_imputed <- data_filtered %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm=TRUE), .)))

# Load model
rf_model_tru_pi <- readRDS("LUAD_Random_Forests/rf_model_final_TRU_vs_proxInflam.rds")

# Predict labels
rf_pred_tru_pi_obj <- predict(rf_model_tru_pi, data_imputed %>% dplyr::select(-all_of(label_col)))

# Extract predictions, named list
rf_pred_tru_pi <- rf_pred_tru_pi_obj$predictions

# Make sure predicted and true labels are factors with same levels
rf_pred_tru_pi <- factor(rf_pred_tru_pi, levels = levels(data_imputed[[label_col]]))
true_labels <- factor(data_imputed[[label_col]], levels = levels(data_imputed[[label_col]]))

# Model evaluation

# Confusion Matrix using true labels
confMat_tru_pi <- confusionMatrix(rf_pred_tru_pi, data_imputed[[label_col]])
# Accuracy: 96.7%

# Saving confusion matrix to revisit if needed, metrics are here
saveRDS(confMat_tru_pi, file = "LUAD_Data/LUAD_Test_Data/Confusion_Matrices/confMat_tru_pi.rds")

# Bootstrapping, check reliability of our results
# Useful for our small test sample size of 31

# Separating input features and labels from test data
X_test <- data_imputed %>% dplyr::select(-all_of(label_col))
y_test <- data_imputed[[label_col]]

# Specify number of bootstrap samples
n_boot <- 1000

# Store metrics per boostrap in a vector
accuracies <- numeric(n_boot)
sensitivities <- numeric(n_boot)
specificities <- numeric(n_boot)
kappas <- numeric(n_boot)

# Assign model here to variable name used by for
rf_model <- rf_model_tru_pi

# Seed for reproducibility
set.seed(145)  

for (i in 1:n_boot) {
  # Sample indices with replacement
  sample_idx <- sample(seq_len(nrow(X_test)), replace = TRUE)
  
  # Create bootstrap sample
  X_boot <- X_test[sample_idx, ]
  y_boot <- y_test[sample_idx]
  
  # Prediction
  pred_boot_obj <- predict(rf_model, X_boot)
  pred_boot <- factor(pred_boot_obj$predictions, levels = levels(y_boot))  # force factor with correct levels
  
  # Evaluation
  cm <- confusionMatrix(pred_boot, y_boot)
  accuracies[i] <- cm$overall["Accuracy"]
  sensitivities[i] <- cm$byClass["Sensitivity"]
  specificities[i] <- cm$byClass["Specificity"]
  kappas[i] <- cm$overall["Kappa"]
}

# Summary of metrics
summary(accuracies)
summary(sensitivities)
summary(specificities)
summary(kappas)

# Overall boostrap results from summary observations for TRU vs proxInflam RF
# Positive class here is proxInflam
# High accuracy: ~95%
# Perfect sensitivity for proxInflam (TP for proxInflam) ~100%
# High specificity (TN, pretty good at detecting TRU) ~93%
# Kappa, agreement beyond chance is hight ~93%

# Metric df to save
bootstrap_results_df_tru_pi <- data.frame(
  accuracy = accuracies,
  sensitivity = sensitivities,
  specificity = specificities,
  kappa = kappas
)

# Save metrics df
write.csv(bootstrap_results_df_tru_pi, "bootstrap_results_TRU_vs_proxInflam.csv", row.names = FALSE)
saveRDS(bootstrap_results_df_tru_pi, file = "bootstrap_results_TRU_vs_proxInflam.rds")

```

#### TRU vs proxProlif Subtype Prediction

```{r model-pred-tru-pp}

# TRU vs proxProlif binary classification
# Load input data

# Set the path to TRU vs proxProlif test data
test_matrix_path <- "LUAD_Test_Data/test_input_matrix_TRU_vs_proxProlif.csv"
# True label col, for model evaluation on unseen data
label_col <- "subtype"

data <- read_csv(test_matrix_path)
str(data)
head(data)

# Remove sample column if present
if ("sample" %in% colnames(data)) {
  data <- data %>% dplyr::select(-sample)
}

# Remove rows where subtype is NA
data <- data %>% filter(!is.na(!!sym(label_col)))
data[[label_col]] <- as.factor(data[[label_col]])

# Remove columns that are all NA
data <- data[, colSums(is.na(data)) < nrow(data)]

# Make sure all features are numeric except label
num_cols <- sapply(data, is.numeric)
num_cols[label_col] <- FALSE
data <- data[, num_cols | names(data) == label_col]

# Drop columns (genes) with >20% NAs, then impute the rest (only 25 samples)
na_threshold <- 0.25 * nrow(data)
data_filtered <- data[, colSums(is.na(data)) < na_threshold]

# Impute remaining NAs with median
data_imputed <- data_filtered %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm=TRUE), .)))

# Load model
rf_model_tru_pp <- readRDS("LUAD_Random_Forests/rf_model_final_TRU_vs_proxProlif.rds")

# Predict labels
rf_pred_tru_pp_obj <- predict(rf_model_tru_pp, data_imputed %>% dplyr::select(-all_of(label_col)))

# Extract predictions, named list
rf_pred_tru_pp <- rf_pred_tru_pp_obj$predictions

# Make sure predicted and true labels are factors with same levels
rf_pred_tru_pp <- factor(rf_pred_tru_pp, levels = levels(data_imputed[[label_col]]))
true_labels <- factor(data_imputed[[label_col]], levels = levels(data_imputed[[label_col]]))

# Model evaluation

# Confusion Matrix using true labels
confMat_tru_pp <- confusionMatrix(rf_pred_tru_pp, data_imputed[[label_col]])
# Accuracy: 96%, pretty good! 
# Saving confusion matrix to revisit if needed, metrics are here
saveRDS(confMat_tru_pp, file = "LUAD_Data/LUAD_Test_Data/Confusion_Matrices/confMat_tru_pp.rds")


# Bootstrapping, check reliability of our results
# Useful for our small test sample size of 25

# Separating input features and labels from test data
X_test <- data_imputed %>% dplyr::select(-all_of(label_col))
y_test <- data_imputed[[label_col]]

# Specify number of bootstrap samples
n_boot <- 1000

# Store metrics per boostrap in a vector
accuracies <- numeric(n_boot)
sensitivities <- numeric(n_boot)
specificities <- numeric(n_boot)
kappas <- numeric(n_boot)

# Assign model here to variable name used by for
rf_model <- rf_model_tru_pp

# Seed for reproducibility
set.seed(145)  

for (i in 1:n_boot) {
  # Sample indices with replacement
  sample_idx <- sample(seq_len(nrow(X_test)), replace = TRUE)
  
  # Create bootstrap sample
  X_boot <- X_test[sample_idx, ]
  y_boot <- y_test[sample_idx]
  
  # Prediction
  pred_boot_obj <- predict(rf_model, X_boot)
  pred_boot <- factor(pred_boot_obj$predictions, levels = levels(y_boot))  # force factor with correct levels
  
  # Evaluation
  cm <- confusionMatrix(pred_boot, y_boot)
  accuracies[i] <- cm$overall["Accuracy"]
  sensitivities[i] <- cm$byClass["Sensitivity"]
  specificities[i] <- cm$byClass["Specificity"]
  kappas[i] <- cm$overall["Kappa"]
}

# Summary of metrics
summary(accuracies)
summary(sensitivities)
summary(specificities)
summary(kappas)

# Overall boostrap results from summary observations for TRU vs proxProlif RF
# Positive class here is proxInflam
# High accuracy: ~96%
# High sensitivity for proxInflam (TP for proxInflam) ~89%
# Perfect specificity (TN, great at detecting TRU) ~100%
# Kappa, agreement beyond chance is hight ~91%

# Metric df to save
bootstrap_results_df_tru_pp <- data.frame(
  accuracy = accuracies,
  sensitivity = sensitivities,
  specificity = specificities,
  kappa = kappas
)

# Save metrics df
write.csv(bootstrap_results_df_tru_pp, "bootstrap_results_TRU_vs_proxProlif.csv", row.names = FALSE)
saveRDS(bootstrap_results_df_tru_pp, file = "bootstrap_results_TRU_vs_proxProlif.rds")
```

### Multi-Class Random Forest Classifier

```{r rf-test-multiclass}

# Load and prepare external test dataset

# Path to the external test feature matrix
test_matrix_path <- "Test Data/LUAD_feature_matrices/test_input_matrix_union_Subtype_Intersected_Genes.csv"

# Load test data
test_data <- read_csv(test_matrix_path)

# Remove 'sample' column if present
if ("sample" %in% colnames(test_data)) test_data <- test_data %>% select(-sample)

# Remove rows with missing subtype labels and convert to factor
test_data <- test_data %>% filter(!is.na(subtype))
test_data$subtype <- as.factor(test_data$subtype)

# Align columns and impute missing values
test_data <- test_data[, colnames(data_imputed)]

# Impute missing values in test data using medians from training data
for (col in setdiff(names(test_data), "subtype")) {
  med <- median(data_imputed[[col]], na.rm = TRUE)
  test_data[[col]][is.na(test_data[[col]])] <- med
}

# Predict class labels using the final trained model
test_pred <- predict(rf_model_final, test_data %>% select(-subtype))

# Create confusion matrix comparing predictions vs true labels
test_conf <- confusionMatrix(test_pred, test_data$subtype)

# Convert confusion matrix to data frame
test_cm_tbl <- as.data.frame(test_conf$table)

# Plot test confusion matrix
p_cm_test <- ggplot(test_cm_tbl, aes(x = Prediction, y = Reference, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), size = 6) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme_minimal() +
  labs(title = "Confusion Matrix (Test set)", x = "Predicted", y = "True")

# Save confusion matrix plot
ggsave(file.path(test_plots_dir, "confusion_matrix_test_union_Subtype_Intersected_Genes.png"),
       plot = p_cm_test, width = 6, height = 5)

# Save confusion matrix table and object
write.csv(test_cm_tbl, file.path(test_conf_dir, "confusion_matrix_test_union_Subtype_Intersected_Genes.csv"), row.names = FALSE)
saveRDS(test_conf, file.path(test_conf_dir, "confusion_matrix_test_union_Subtype_Intersected_Genes.rds"))

# Save raw test predictions
write.csv(
  data.frame(True = test_data$subtype, Pred = test_pred),
  file.path(test_pred_dir, "rf_test_predictions_union_Subtype_Intersected_Genes.csv"),
  row.names = FALSE
)

```

Bootstrapping to make up for the small sample size in the test data.

```{r rf-test-muliclass-bootstrap}

n_boot <- 1000
accs <- numeric(n_boot)
set.seed(123)
for (i in 1:n_boot) {
  idx <- sample(seq_len(nrow(test_data)), replace = TRUE)
  x_boot <- test_data[idx, ] %>% select(-subtype)
  y_boot <- test_data$subtype[idx]
  pred_boot <- predict(rf_model_final, x_boot)
  accs[i] <- mean(pred_boot == y_boot)
}
boot_df <- data.frame(Accuracy=accs)
write.csv(boot_df, file.path(bootstrap_dir, "bootstrap_results_test_union_Subtype_Intersected_Genes.csv"), row.names = FALSE)
saveRDS(boot_df, file.path(bootstrap_dir, "bootstrap_results_test_union_Subtype_Intersected_Genes.rds"))

# Histogram plot (Test)
p_boot_hist <- ggplot(boot_df, aes(x=Accuracy)) +
  geom_histogram(bins=30, fill="steelblue", color="black") +
  labs(title="Bootstrap Accuracy Distribution (Test set)", x="Accuracy", y="Count") +
  theme_minimal()

ggsave(file.path(test_plots_dir, "bootstrap_hist_accuracy_test_union_Subtype_Intersected_Genes.png"), plot=p_boot_hist, width=6, height=4)

# Boxplot (Test)
boot_long <- boot_df %>% pivot_longer(everything(), names_to="Metric", values_to="Value")
p_box <- ggplot(boot_long, aes(x=Metric, y=Value, fill=Metric)) +
  geom_boxplot(alpha=0.7) + theme_minimal() +
  scale_fill_brewer(palette="Set2") +
  labs(title="Bootstrap Metrics Boxplot (Test set)")

ggsave(file.path(test_plots_dir, "bootstrap_boxplot_test_union_Subtype_Intersected_Genes.png"), plot=p_box, width=6, height=4)

cat(sprintf("Bootstrap mean accuracy: %.3f, 95%% CI: [%.3f, %.3f]\n",
            mean(accs), quantile(accs,0.025), quantile(accs,0.975)))

```

In this notebook we have processed our test set, produced a normalised and scaled test feature matrix, used it as input for our trained models and evaluated model accuracy on unseen data.
