# Biological Annotation and Interpretation

## Project Overview

This notebook performs functional annotation and biological interpretation of genes identified through differential analyses and machine learning feature selection.

The goal is to identify biological pathways, molecular processes, and disease associations underlying LUAD subtype classification.

## Methods

- Gene identifier harmonization
- GO enrichment analysis
- KEGG pathway enrichment
- Reactome enrichment
- Disease Ontology analysis
- Feature overlap analysis

## Outputs

- Enriched pathways
- Functional annotations
- Gene overlap summaries
- Biological interpretation figures

In the annotation part of the project we want to link machine learning features and differentially regulated genes to biological meaning.

In order to do so, we:

1)  Convert gene IDs to a standard format (Entrez) to make them usable in enrichment tools.
2)  Use databases like GO, KEGG, Reactome, and Disease Ontology to find biological processes, pathways, and diseases related to these genes.
3)  Make plots that show the most enriched biological functions.
4)  Compare genes from different sources (DEG, DMP, Random Forest, SVM) to find overlaps.
5)  Save all plots and overlapping genes for further interpretation.

## Setup

```{r output-directory, message=FALSE, warning=FALSE}
results_base_dir <- "/Users/mac/Desktop/results_LUAD_5"
dir.create(results_base_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load required libraries

```{r libraries, message=FALSE, warning=FALSE}

# Install Bioconductor if needed 
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")

# Bioconductor packages
# BiocManager::install(c(
#   "clusterProfiler", "ReactomePA", "DOSE", "org.Hs.eg.db", "AnnotationDbi"
# ))

# CRAN packages
# install.packages(c("tidyverse", "ggvenn", "ggplot2", "dplyr", "readr"))

library(tidyverse)      # data manipulation and visualization 
library(AnnotationDbi)  # base tools for annotation databases
library(org.Hs.eg.db)   # human gene annotation database 
library(clusterProfiler) # functional enrichment analysis (GO, KEGG, DO)
library(enrichplot)     # visualization of enrichment results
library(ReactomePA)     # pathway enrichment using Reactome database
library(DOSE)           # disease ontology enrichment analysis
library(ggvenn)         # Venn diagrams based on ggplot2

```

## Define Comparisons List

```{r comparisons-list, message=FALSE, warning=FALSE}

# This section defines a list of all pairwise comparisons to be analyzed.
# Each comparison includes:
# - name: a label for the comparison (used for folder names and output)
# - deg: path to CSV file with differentially expressed genes (DEGs)
# - dmp: path to CSV file with differentially methylated positions (DMPs)
# - rf: path to CSV file with top features from the Random Forest classifier
# - svm: path to CSV file with top features from the Support Vector Machine classifier
comparisons <- list(
  list(
    name = "TRU_vs_proxInflam",
    deg  = "/Users/mac/Desktop/data_science_project/luad_pi_vs_tru/luad_pi_vs_tru_prox_-inflam_vs_TRU_DEG_filtered.csv",
    dmp  = "/Users/mac/Downloads/DMP_TRU_vs_proxInflam.csv",
    rf   = "/Users/mac/Downloads/important_features_TRU_vs_proxInflam_RF_new.csv",
    svm  = "/Users/mac/Downloads/feature_importance_TRU_vs_proxInflam_SVM_new.csv"
  ),
  list(
    name = "TRU_vs_proxProlif",
    deg  = "/Users/mac/Desktop/data_science_project/luad_pp_vs_tru/luad_pp_vs_tru_prox_-prolif__vs_TRU_DEG_filtered.csv",
    dmp  = "/Users/mac/Downloads/DMP_TRU_vs_proxProlif.csv",
    rf   = "/Users/mac/Downloads/important_features_TRU_vs_proxProlif_RF_new.csv",
    svm  = "/Users/mac/Downloads/feature_importance_TRU_vs_proxProlif_SVM_new.csv"
  ),
  list(
    name = "proxInflam_vs_proxProlif",
    deg  = "/Users/mac/Desktop/data_science_project/luad_pi_vs_pp/luad_pi_vs_pp_prox_-inflam_vs_prox_-prolif__DEG_filtered.csv",
    dmp  = "/Users/mac/Downloads/DMP_proxProlif_vs_proxInflam.csv",
    rf   = "/Users/mac/Downloads/important_features_proxProlif_vs_proxInflam_RF_new.csv",
    svm  = "/Users/mac/Downloads/feature_importance_proxProlif_vs_proxInflam_SVM_new.csv"
  )
)

```

## Save Plot Function

```{r save-plot, message=FALSE, warning=FALSE}

# This function will hepl to save any ggplot or compatible plot as a high-resolution PNG file
# - filename: output file path
# - plot_expr: the plot expression or object to be printed

save_plot <- function(filename, plot_expr, width = 10, height = 6) {
  png(filename, width = width, height = height, units = "in", res = 300)
  print(plot_expr)
  dev.off()
}

```

## Safe Gene ID Mapping

```{r safe-map, message=FALSE, warning=FALSE}

# Function to safely convert gene identifiers between formats
# - genes: vector of input gene IDs
# - from: source ID type ("ENSEMBL", "SYMBOL")
# - to: target ID type ("ENTREZID")
# It filters only valid IDs and removes NA values after mapping.

safe_map <- function(genes, from, to) {
  valid_genes <- genes[genes %in% keys(org.Hs.eg.db, keytype = from)]
  if (length(valid_genes) == 0) return(character(0))
  res <- mapIds(org.Hs.eg.db, keys = valid_genes, column = to, keytype = from, multiVals = "first")
  na.omit(res)
}

```

## Annotation

```{r annotation-function, message=FALSE, warning=FALSE}

# Here we will perform enrichment analysis for different data types and save the plots and overlap statistics for each comparison

run_annotation_analysis <- function(deg_df, dmp_df, rf_df, svm_df, out_dir, comparison_name) {
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# DEG analysis
# Cleaning gene identifiers (by removing version suffixes), droping duplicates, converting ENSEMBL to ENTREZID and performing GO, KEGG, Reactome, DO enrichment
  
  deg_df$gene <- gsub("\\..*", "", deg_df$gene)
  deg_df <- deg_df[!duplicated(deg_df$gene), ]
  entrez_deg <- safe_map(deg_df$gene, "ENSEMBL", "ENTREZID")
  if (length(entrez_deg) >= 10) {
    save_plot(file.path(out_dir, "DEG_GO.png"), dotplot(enrichGO(entrez_deg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID", ont = "BP", pvalueCutoff = 0.05, readable = TRUE), showCategory = 10))
    save_plot(file.path(out_dir, "DEG_Reactome.png"), dotplot(enrichPathway(entrez_deg, organism = "human", pvalueCutoff = 0.05, readable = TRUE), showCategory = 10))
    save_plot(file.path(out_dir, "DEG_KEGG.png"), dotplot(enrichKEGG(entrez_deg, organism = "hsa", pvalueCutoff = 0.05), showCategory = 10))
    save_plot(file.path(out_dir, "DEG_DO.png"), barplot(enrichDO(entrez_deg, ont = "DO", pvalueCutoff = 0.05, readable = TRUE), showCategory = 10))
  }
  
# DMP analysis
# Filtering and deduplicating SYMBOLs from DMPs, maping to ENTREZID and runing functional enrichment
  
  dmp_df <- dmp_df[!is.na(dmp_df$gene_HGNC) & dmp_df$gene_HGNC != "", ]
  dmp_df <- dmp_df[!duplicated(dmp_df$gene_HGNC), ]
  entrez_dmp <- safe_map(dmp_df$gene_HGNC, "SYMBOL", "ENTREZID")
  if (length(entrez_dmp) >= 10) {
    save_plot(file.path(out_dir, "DMP_GO.png"), dotplot(enrichGO(entrez_dmp, OrgDb = org.Hs.eg.db, keyType = "ENTREZID", ont = "BP", pvalueCutoff = 0.05, readable = TRUE), showCategory = 10))
    save_plot(file.path(out_dir, "DMP_Reactome.png"), dotplot(enrichPathway(entrez_dmp, organism = "human", pvalueCutoff = 0.05, readable = TRUE), showCategory = 10))
    save_plot(file.path(out_dir, "DMP_KEGG.png"), dotplot(enrichKEGG(entrez_dmp, organism = "hsa", pvalueCutoff = 0.05), showCategory = 10))
    save_plot(file.path(out_dir, "DMP_DO.png"), barplot(enrichDO(entrez_dmp, ont = "DO", pvalueCutoff = 0.05, readable = TRUE), showCategory = 10))
  }

# Random Forest analysis
# Extracting gene names from feature column, scaling importance, maping to ENTREZID and runing enrichment
  
  rf_df <- rf_df %>%
    mutate(gene = gsub("^(expr_|meth_)", "", feature)) %>%
    mutate(scaled_importance = importance / max(importance))
  entrez_rf <- safe_map(rf_df$gene, "SYMBOL", "ENTREZID")
  if (length(entrez_rf) >= 10) {
    save_plot(file.path(out_dir, "RF_GO.png"), dotplot(enrichGO(entrez_rf, OrgDb = org.Hs.eg.db, keyType = "ENTREZID", ont = "BP", pvalueCutoff = 0.1, readable = TRUE), showCategory = 10))
    save_plot(file.path(out_dir, "RF_Reactome.png"), dotplot(enrichPathway(entrez_rf, organism = "human", pvalueCutoff = 0.05, readable = TRUE), showCategory = 10))
    save_plot(file.path(out_dir, "RF_KEGG.png"), dotplot(enrichKEGG(entrez_rf, organism = "hsa", pvalueCutoff = 0.05), showCategory = 10))
    save_plot(file.path(out_dir, "RF_DO.png"), barplot(enrichDO(entrez_rf, ont = "DO", pvalueCutoff = 0.05, readable = TRUE), showCategory = 10))
  }
  
# SVM analysis
# Same idea here: extracting gene names from feature column, scaling importance, maping to ENTREZID and runing enrichment

  svm_df <- svm_df %>%
    mutate(gene = gsub("^(expr_|meth_)", "", Feature)) %>%
    mutate(scaled_importance = Importance / max(Importance))
  entrez_svm <- safe_map(svm_df$gene, "SYMBOL", "ENTREZID")
  if (length(entrez_svm) >= 10) {
    save_plot(file.path(out_dir, "SVM_GO.png"), dotplot(enrichGO(entrez_svm, OrgDb = org.Hs.eg.db, keyType = "ENTREZID", ont = "BP", pvalueCutoff = 0.1, readable = TRUE), showCategory = 10))
    save_plot(file.path(out_dir, "SVM_Reactome.png"), dotplot(enrichPathway(entrez_svm, organism = "human", pvalueCutoff = 0.05, readable = TRUE), showCategory = 10))
    save_plot(file.path(out_dir, "SVM_KEGG.png"), dotplot(enrichKEGG(entrez_svm, organism = "hsa", pvalueCutoff = 0.05), showCategory = 10))
    save_plot(file.path(out_dir, "SVM_DO.png"), barplot(enrichDO(entrez_svm, ont = "DO", pvalueCutoff = 0.05, readable = TRUE), showCategory = 10))
  }

# Overlap and Venn diagram
# Collecting all unique gene names and plotting Venn diagram to chech wheather features are overlaping
  
  overlap_genes <- list(
    DEG = unique(deg_df$gene),
    DMP = unique(dmp_df$gene_HGNC),
    RF  = unique(rf_df$gene),
    SVM = unique(svm_df$gene)
  )

  venn_plot <- ggvenn(overlap_genes, fill_color = c("red", "blue", "green", "purple"))
  save_plot(file.path(out_dir, "Venn_All.png"), venn_plot, 8, 8)
  
# Saving intersecting genes to text files
  writeLines(Reduce(intersect, overlap_genes), file.path(out_dir, "Common_All.txt"))
  writeLines(intersect(overlap_genes$RF, overlap_genes$SVM), file.path(out_dir, "Common_RF_SVM.txt"))
  writeLines(intersect(overlap_genes$DEG, overlap_genes$RF), file.path(out_dir, "Common_DEG_RF.txt"))
  writeLines(intersect(overlap_genes$DMP, overlap_genes$DEG), file.path(out_dir, "Common_DMP_DEG.txt"))
}

```

## Run Pairwise Annotations 

```{r run-annotations, message=FALSE, warning=FALSE}

# Here we make a loop that iterates over all pairwise comparisons defined in the `comparisons` list
# For each comparison it:
# - Reads DEG, DMP, RF, and SVM CSV files into data frames
# - Creates a dedicated output folder
# - Runs enrichment analysis and saves all plots and overlap files

# The main function being called is `run_annotation_analysis()`

for (comp in comparisons) {
  run_annotation_analysis(
    deg_df = read.csv(comp$deg),
    dmp_df = read.csv(comp$dmp),
    rf_df  = read.csv(comp$rf),
    svm_df = read.csv(comp$svm),
    out_dir = file.path(results_base_dir, comp$name),
    comparison_name = comp$name
  )
}

```

## Multiclass RF and SVM Analysis

```{r multiclass-analysis, message=FALSE, warning=FALSE}

# In this section we analyze features selected by RF and SVM models across all subtypes (multiclass).
# We want to identify overlapping genes, perfors functional enrichment and visualization of shared importance

# Read and preprocess Random Forest results 
# Removing prefixes (expr_/meth_) and scale importance

multi_rf <- read.csv("/Users/mac/Downloads/important_features_union_Subtype_Intersected_Genes_RF_multi.csv") %>%
  mutate(gene = gsub("^(expr_|meth_)", "", feature)) %>%
  mutate(scaled_importance = MeanDecreaseAccuracy / max(MeanDecreaseAccuracy))

# Read and preprocess SVM results 
# Removing prefixes (expr_/meth_) and scale importance

multi_svm <- read.csv("/Users/mac/Downloads/feature_importance_union_Subtype_Intersected_Genes_SVM_multi.csv") %>%
  mutate(gene = gsub("^(expr_|meth_)", "", Feature)) %>%
  mutate(scaled_importance = Importance / max(Importance))

# Identify overlapping genes between RF and SVM 

multi_overlap_genes <- intersect(multi_rf$gene, multi_svm$gene)
writeLines(multi_overlap_genes, file.path(results_base_dir, "Multiclass_Common_RF_SVM.txt"))

# Draw a Venn diagram of shared features
venn_multi <- ggvenn(list(Multi_RF = multi_rf$gene, Multi_SVM = multi_svm$gene), fill_color = c("orange", "skyblue"))
save_plot(file.path(results_base_dir, "Venn_Multiclass_RF_SVM.png"), venn_multi, 7, 7)

# Enrichment analysis on overlapping genes
entrez_multi <- safe_map(multi_overlap_genes, "SYMBOL", "ENTREZID")
if (length(entrez_multi) >= 10) {
  save_plot(file.path(results_base_dir, "Multiclass_Overlap_GO.png"), dotplot(enrichGO(entrez_multi, OrgDb = org.Hs.eg.db, keyType = "ENTREZID", ont = "BP", pvalueCutoff = 0.1, readable = TRUE), showCategory = 10))
  save_plot(file.path(results_base_dir, "Multiclass_Overlap_Reactome.png"), dotplot(enrichPathway(entrez_multi, organism = "human", pvalueCutoff = 0.1, readable = TRUE), showCategory = 10))
  save_plot(file.path(results_base_dir, "Multiclass_Overlap_KEGG.png"), dotplot(enrichKEGG(entrez_multi, organism = "hsa", pvalueCutoff = 0.05), showCategory = 10))
}

# Combine and rank gene importance from both models 

importance_df <- full_join(
  multi_rf %>% filter(gene %in% multi_overlap_genes) %>% select(gene, MeanDecreaseAccuracy),
  multi_svm %>% filter(gene %in% multi_overlap_genes) %>% select(gene, Importance),
  by = "gene"
) %>% mutate(total = MeanDecreaseAccuracy + Importance) %>%
  arrange(desc(total))

write.csv(importance_df, file.path(results_base_dir, "Multiclass_Overlap_RF_SVM_Importance.csv"), row.names = FALSE)

# Select top overlapping genes and visualize their combined importance 

top_overlap <- importance_df %>% slice_max(order_by = total, n = 20)

bar_plot <- ggplot(top_overlap, aes(x = reorder(gene, total), y = total)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top Overlapping Genes by RF + SVM Importance",
       y = "Combined Importance", x = "Gene") +
  theme_minimal()

save_plot(file.path(results_base_dir, "Multiclass_Overlap_Importance_Barplot.png"), bar_plot, 7, 6)

```
